---
title: "Tirzepatide vs Semaglutide 2.4 mg: Persistence, Utilization and Costs"
subtitle: "Annotated analysis of a synthetic commercial claims extract (target trial emulation)"
author: "Maxwell K. Nyarko"
date: "October 2026"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: false
    df_print: paged
---

```{r knitr-options, include = FALSE}
# Run every chunk from the project folder; show code and results; do not stop on warnings
knitr::opts_knit$set(root.dir = here::here())
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 8, fig.height = 5)
```

# How to use this file

1. Open `HEOR_Tirzepatide_Project.Rproj` in RStudio (this sets the working folder).
2. Open this file and either click **Knit** (runs everything and makes an HTML report) or run the chunks one at a time from top to bottom with the green arrow on each chunk (Ctrl/Cmd + Shift + Enter).
3. The first chunk installs any missing packages, which can take a few minutes the first time. A full run takes about 5-10 minutes; the bootstrap in Step 9.6 is the slowest part.

Every step below has the same layout: **Step**, **What** (each key action and the code line that does it), **Why**, then the R code. Each step saves its results to `04_data_derived/` or `06_output/`, so you can stop and restart at any step after the setup chunk.

**The data are fully synthetic.** Results show how the methods work and are not evidence about either drug.

The same code is also saved as numbered scripts in `05_code/` (00 to 10); this file is assembled from those scripts by `05_code/98_build_rmd.R`.

# Step 0. Set up R

**What**

- Install and load the packages for data handling, weighting, models and figures: `install.packages(to_install)` and `lapply(pkgs, library, ...)`.
- Make `select()` refer to dplyr (MASS also has a `select()`): `select <- dplyr::select`.

**Why.** Every later step uses these packages. Loading them once at the start makes the run reproducible and avoids errors halfway through.

```{r setup-packages}
pkgs <- c(
  # data handling
  "here", "data.table", "dplyr", "tidyr", "stringr", "lubridate", "readr", "janitor",
  # descriptive tables
  "gtsummary", "tableone", "flextable", "officer",
  # confounding adjustment and balance
  "WeightIt", "MatchIt", "cobalt",
  # models
  "survival", "survminer", "sandwich", "lmtest", "broom", "MASS", "pscl", "marginaleffects",
  # sensitivity analysis
  "EValue",
  # DAG
  "dagitty", "ggdag",
  # figures
  "ggplot2", "patchwork", "scales"
)
to_install <- setdiff(pkgs, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install, repos = "https://cloud.r-project.org")
suppressPackageStartupMessages(invisible(lapply(pkgs, library, character.only = TRUE)))
# MASS masks dplyr::select; make sure select() means dplyr::select()
select <- dplyr::select
```

**What**

- Define folder paths with `here::here()` so the code works on any computer.
- Store the study calendar from the protocol: identification window (`index_start`, `index_end`), 365-day baseline and follow-up, the 60-day persistence gap (`gap_days`), and the 30-day allowed enrollment gap.
- Fix the random seed with `set.seed(20261007)` so the bootstrap gives the same answer each time.

**Why.** Keeping all design constants in one place means a design change is made once, and readers can check the code against the SAP.

```{r setup-paths}
dir_raw     <- here::here("02_data_raw")
dir_ref     <- here::here("03_reference")
dir_derived <- here::here("04_data_derived")
dir_tables  <- here::here("06_output", "tables")
dir_figures <- here::here("06_output", "figures")
for (d in c(dir_derived, dir_tables, dir_figures)) dir.create(d, showWarnings = FALSE, recursive = TRUE)

# ---- study calendar (from the protocol) --------------------------------------
data_start    <- as.Date("2023-01-01")
data_end      <- as.Date("2026-06-30")
index_start   <- as.Date("2024-01-01")
index_end     <- as.Date("2025-06-30")
baseline_days <- 365   # baseline = index - 365 to index - 1
followup_days <- 365   # follow-up = index to index + 364
gap_days      <- 60    # primary persistence grace period (SAP section 6.1)
enroll_gap_allowed <- 30  # enrollment gaps of <= 30 days are bridged (SAP section 4)

set.seed(20261007)
theme_set(theme_minimal(base_size = 11))
col_drug <- c("Tirzepatide" = "#1F6FB4", "Semaglutide 2.4 mg" = "#D9822B")
message("Setup complete. Raw data in: ", dir_raw)
```

# Step 1. Study design: the DAG (workflow step 3)

**What**

- Write the assumed causal structure as a DAG with `dagitty::dagitty()`: one for persistence (`dag_persist`) and one for costs (`dag_cost`).
- Mark the exposure (`Drug`) and outcome, and list unmeasured nodes (`SES_U`, `Preference_U`).

**Why.** The DAG makes our assumptions explicit and tells us which variables to adjust for (confounders) and which to leave alone (mediators such as GI side effects or the drug's copay).

```{r dag-define}
# Nodes measured in claims are written in CamelCase; unmeasured nodes end in "_U".
dag_persist <- dagitty::dagitty('dag {
  Drug [exposure]
  Persistence [outcome]
  Age -> {Drug Persistence ObesitySeverity Comorbidity BaselineUse}
  Sex -> {Drug Persistence Comorbidity}
  Region -> {Drug Persistence PlanType}
  PlanType -> {Drug Persistence OOPCost}
  CalendarTime -> {Drug Persistence}
  ObesitySeverity -> {Drug Persistence Comorbidity}
  Comorbidity -> {Drug Persistence BaselineUse}
  PriorAOM -> {Drug Persistence}
  BaselineUse -> {Drug Persistence}
  Endocrinologist -> {Drug Persistence}
  SES_U -> {PlanType Region Persistence}
  Preference_U -> {Drug}
  Drug -> {GI_AE WeightLoss OOPCost Persistence}
  GI_AE -> Persistence
  WeightLoss -> Persistence
  OOPCost -> Persistence
  Comorbidity -> Endocrinologist
  ObesitySeverity -> PriorAOM
}')

# Cost outcome DAG: persistence becomes a mediator between drug and costs
dag_cost <- dagitty::dagitty('dag {
  Drug [exposure]
  Costs [outcome]
  Age -> {Drug Costs ObesitySeverity Comorbidity BaselineUse}
  Sex -> {Drug Costs Comorbidity}
  Region -> {Drug Costs PlanType}
  PlanType -> {Drug Costs}
  CalendarTime -> {Drug Costs}
  ObesitySeverity -> {Drug Costs Comorbidity}
  Comorbidity -> {Drug Costs BaselineUse}
  PriorAOM -> {Drug Costs}
  BaselineUse -> {Drug Costs}
  Endocrinologist -> {Drug Costs}
  SES_U -> {PlanType Region Costs}
  Preference_U -> {Drug}
  Drug -> {Persistence GI_AE Costs}
  Persistence -> Costs
  GI_AE -> {Persistence Costs}
  Comorbidity -> Endocrinologist
  ObesitySeverity -> PriorAOM
}')
```

**What**

- Ask dagitty for the minimal sufficient adjustment set: `adjustmentSets(dag_persist, type = "minimal")`.
- List the descendants of `Drug`, which must not be adjusted for: `descendants(dag_persist, "Drug")`.

**Why.** This turns the DAG into a concrete covariate list for the propensity score model (Step 8).

```{r dag-adjust}
# Minimal sufficient adjustment sets (unmeasured nodes cannot be used)
latent <- c("SES_U", "Preference_U")
dagitty::latents(dag_persist) <- latent
dagitty::latents(dag_cost)    <- latent
adj_persist <- dagitty::adjustmentSets(dag_persist, type = "minimal")
adj_cost    <- dagitty::adjustmentSets(dag_cost, type = "minimal")
print(adj_persist); print(adj_cost)
cat("Mediators NOT to adjust for (descendants of Drug):",
    paste(setdiff(dagitty::descendants(dag_persist, "Drug"), c("Drug", "Persistence")), collapse = ", "), "\n")
writeLines(c("Persistence DAG - minimal adjustment set(s):", capture.output(print(adj_persist)),
             "", "Cost DAG - minimal adjustment set(s):", capture.output(print(adj_cost))),
           file.path(dir_tables, "dag_adjustment_sets.txt"))
```

**What.** Plot both DAGs with nodes coloured by role (`plot_dag()`) and save them to `06_output/figures/`.

**Why.** A picture of the assumptions goes into the DAG document and the paper.

```{r dag-plot, fig.height = 7}
node_role <- function(dag, outcome) {
  tidy <- ggdag::tidy_dagitty(dag, layout = "sugiyama", seed = 3)
  tidy$data <- tidy$data |>
    mutate(role = case_when(
      name == "Drug" ~ "Exposure",
      name == outcome ~ "Outcome",
      grepl("_U$", name) ~ "Unmeasured",
      name %in% c("GI_AE", "WeightLoss", "OOPCost", "Persistence") ~ "Mediator (do not adjust)",
      TRUE ~ "Measured confounder (adjusted)"))
  tidy
}
plot_dag <- function(dag, outcome, title) {
  ggplot(node_role(dag, outcome), aes(x = x, y = y, xend = xend, yend = yend)) +
    ggdag::geom_dag_edges(edge_colour = "grey55", edge_width = 0.35) +
    ggdag::geom_dag_point(aes(colour = role), size = 19) +
    ggdag::geom_dag_text(colour = "black", size = 2.6) +
    scale_colour_manual(values = c("Exposure" = "#7FB3E0", "Outcome" = "#F2B880",
                                   "Unmeasured" = "#D9D9D9", "Mediator (do not adjust)" = "#E6A0A0",
                                   "Measured confounder (adjusted)" = "#A8D5A2"), name = NULL) +
    ggdag::theme_dag() + theme(legend.position = "bottom") +
    guides(colour = guide_legend(nrow = 2)) +
    labs(title = title)
}
p_dag1 <- plot_dag(dag_persist, "Persistence", "DAG 1. Tirzepatide vs semaglutide 2.4 mg and 12-month persistence")
p_dag2 <- plot_dag(dag_cost, "Costs", "DAG 2. Tirzepatide vs semaglutide 2.4 mg and 12-month costs / utilization")
ggsave(file.path(dir_figures, "fig0_dag_persistence.png"), p_dag1, width = 11, height = 7.5, dpi = 200, bg = "white")
ggsave(file.path(dir_figures, "fig0_dag_costs.png"), p_dag2, width = 11, height = 7.5, dpi = 200, bg = "white")
```

# Step 2. Import the raw extract (workflow step 7 begins)

**What**

- Declare column types so all codes are read as text (`rx_types`, `med_types`; e.g. `ndc = col_character()`).
- Read and stack the yearly gzip files with `read_stack()`.
- Read the members, enrollment and reference tables with `readr::read_csv()`.

**Why.** Reading NDC, ICD-10, CPT and place-of-service codes as numbers would drop leading zeros and silently break code matching. Nothing is cleaned here, so the raw data stay exactly as delivered.

```{r import-raw}

# Codes must be read as character so leading zeros are not lost on import.
rx_types <- cols(
  rx_claim_id = col_character(), member_id = col_character(), fill_date = col_date(),
  ndc = col_character(), days_supply = col_integer(), quantity = col_double(),
  refill_number = col_integer(), claim_status = col_character(),
  pharmacy_channel = col_character(), allowed_amt = col_double(),
  paid_amt = col_double(), member_oop = col_double()
)
med_types <- cols(
  claim_id = col_character(), line_num = col_integer(), member_id = col_character(),
  claim_type = col_character(), service_from_date = col_date(), service_to_date = col_date(),
  admit_date = col_date(), discharge_date = col_date(), place_of_service = col_character(),
  revenue_code = col_character(), proc_code = col_character(),
  dx1 = col_character(), dx2 = col_character(), dx3 = col_character(),
  dx4 = col_character(), dx5 = col_character(), prov_specialty = col_character(),
  allowed_amt = col_double(), paid_amt = col_double()
)

read_stack <- function(pattern, types) {
  files <- list.files(dir_raw, pattern = pattern, full.names = TRUE)
  message("Reading: ", paste(basename(files), collapse = ", "))
  dplyr::bind_rows(lapply(files, readr::read_csv, col_types = types, progress = FALSE))
}

members    <- readr::read_csv(file.path(dir_raw, "members.csv"),
                              col_types = cols(.default = col_character(), birth_year = col_integer()))
enrollment <- readr::read_csv(file.path(dir_raw, "enrollment.csv"),
                              col_types = cols(.default = col_character(),
                                               eligibility_start = col_date(), eligibility_end = col_date()))
pharmacy   <- read_stack("^pharmacy_claims_\\d{4}\\.csv\\.gz$", rx_types)
medical    <- read_stack("^medical_claims_\\d{4}\\.csv\\.gz$", med_types)

# reference tables
ndc_lookup <- readr::read_csv(file.path(dir_ref, "ndc_product_lookup.csv"), col_types = cols(.default = col_character()))
icd_lookup <- readr::read_csv(file.path(dir_ref, "icd10cm_codes.csv"), col_types = cols(.default = col_character()))
```

**What.** Print row and column counts and parsing problems for each table, then save R copies with `saveRDS()`.

**Why.** A structural check right after import catches truncated files or type problems before they cause trouble later.

```{r import-check}
# quick structural check (row counts, parsing problems)
for (nm in c("members", "enrollment", "pharmacy", "medical")) {
  x <- get(nm)
  message(sprintf("%-11s rows: %s  cols: %s  parse problems: %s",
                  nm, format(nrow(x), big.mark = ","), ncol(x), nrow(readr::problems(x))))
}

saveRDS(members,    file.path(dir_derived, "raw_members.rds"))
saveRDS(enrollment, file.path(dir_derived, "raw_enrollment.rds"))
saveRDS(pharmacy,   file.path(dir_derived, "raw_pharmacy.rds"))
saveRDS(medical,    file.path(dir_derived, "raw_medical.rds"))
message("Raw tables saved to 04_data_derived/")
```

# Step 3. Data quality review

**What**

- Reload the raw tables (`readRDS()`).
- Create a log with one row per check, using `log_check(table, check, n_affected, decision)`.

**Why.** Vendor claims always contain errors. Finding them before cleaning, and writing down a decision for each, makes the cleaning transparent and reproducible (workflow step 7: "log every data-cleaning decision").

```{r dq-load}
members    <- readRDS(file.path(dir_derived, "raw_members.rds"))
enrollment <- readRDS(file.path(dir_derived, "raw_enrollment.rds"))
pharmacy   <- readRDS(file.path(dir_derived, "raw_pharmacy.rds"))
medical    <- readRDS(file.path(dir_derived, "raw_medical.rds"))
ndc_lookup <- readr::read_csv(file.path(dir_ref, "ndc_product_lookup.csv"),
                              col_types = cols(.default = col_character()))

# Each check adds one row: table, check, number of rows affected, decision taken
dq <- list()
log_check <- function(table, check, n, decision) {
  dq[[length(dq) + 1]] <<- tibble(table = table, check = check, n_affected = n, decision = decision)
}
```

**What.** Members: count exact duplicates (`duplicated()`), IDs with conflicting demographics (`count(member_id) |> filter(n > 1)`), missing birth year or state, and unknown sex.

**Why.** A member ID should identify one person. Conflicting records mean we cannot know the person's sex or age.

```{r dq-members}
dup_ids     <- members$member_id[duplicated(members$member_id)]
exact_dups  <- sum(duplicated(data.table::as.data.table(members)))
conflicting <- members |> distinct() |> count(member_id) |> filter(n > 1) |> nrow()
log_check("members", "Exact duplicate rows", exact_dups, "Drop exact duplicates")
log_check("members", "Member IDs with conflicting demographics (e.g., different sex)", conflicting,
          "Cannot tell which record is right: exclude these members from the cohort")
log_check("members", "Missing birth year", sum(is.na(members$birth_year)),
          "Age cannot be computed: excluded at the age criterion")
log_check("members", "Sex unknown (U)", sum(members$sex == "U", na.rm = TRUE),
          "Excluded (complete-case on sex; < 1% of members)")
log_check("members", "Missing state", sum(is.na(members$state)), "Region set to 'Unknown' category")
log_check("members", "Birth year implies age < 18 in 2024", sum(members$birth_year > 2006, na.rm = TRUE),
          "Removed by the age 18-64 criterion")
```

**What.** Enrollment: count duplicate segments, overlapping segments (`eligibility_start <= prev_end`), gaps (`eligibility_start > prev_end + 1`), and segments without drug or medical coverage.

**Why.** Continuous enrollment is an eligibility criterion, and pharmacy claims exist only while drug coverage is active.

```{r dq-enrollment}
enr_sorted <- enrollment |> distinct() |> arrange(member_id, eligibility_start) |>
  group_by(member_id) |> mutate(prev_end = lag(eligibility_end)) |> ungroup()
log_check("enrollment", "Exact duplicate segments", sum(duplicated(data.table::as.data.table(enrollment))), "Drop exact duplicates")
log_check("enrollment", "Overlapping segments for the same member",
          sum(enr_sorted$eligibility_start <= enr_sorted$prev_end, na.rm = TRUE),
          "Collapse overlapping segments into one continuous span")
log_check("enrollment", "Gaps between segments (> 1 day)",
          sum(enr_sorted$eligibility_start > enr_sorted$prev_end + 1, na.rm = TRUE),
          "Gaps <= 30 days bridged; longer gaps break continuous enrollment")
log_check("enrollment", "Segments without pharmacy coverage (rx_coverage = N)",
          sum(enrollment$rx_coverage != "Y"), "Not counted as covered time (pharmacy claims not captured)")
log_check("enrollment", "Segments without medical coverage", sum(enrollment$medical_coverage != "Y"),
          "Not counted as covered time")
```

**What.** Pharmacy: count the three NDC formats (`nchar(pharmacy$ndc)`), duplicates, reversals (`claim_status == "R"`), missing allowed amounts, and implausible days supply (< 3 or > 90).

**Why.** NDCs must match the lookup to identify the study drugs; reversals are claims that were never dispensed; days supply drives persistence and PDC.

```{r dq-pharmacy}
log_check("pharmacy", "NDC hyphenated 5-4-2 format (13 characters)", sum(nchar(pharmacy$ndc) == 13),
          "Remove hyphens")
log_check("pharmacy", "NDC hyphenated 4-4-2 format (12 characters, labeler lost a leading zero)",
          sum(nchar(pharmacy$ndc) == 12), "Remove hyphens and pad labeler to 5 digits")
log_check("pharmacy", "NDC stored as number (leading zeros lost; < 11 digits)",
          sum(!grepl("-", pharmacy$ndc) & nchar(pharmacy$ndc) < 11), "Left-pad with zeros to 11 digits")
log_check("pharmacy", "Exact duplicate rows", sum(duplicated(data.table::as.data.table(pharmacy))), "Drop exact duplicates")
log_check("pharmacy", "Reversed claims (claim_status = R)", sum(pharmacy$claim_status == "R"),
          "Drop the reversal and the paid claim it reverses (same rx_claim_id)")
paid <- pharmacy |> filter(claim_status == "P")
log_check("pharmacy", "Same paid rx_claim_id appearing more than once (format-only differences)",
          sum(duplicated(paid$rx_claim_id)) - sum(duplicated(data.table::as.data.table(paid))), "Keep one row per rx_claim_id")
log_check("pharmacy", "Missing allowed amount", sum(is.na(pharmacy$allowed_amt)),
          "Impute as paid_amt + member_oop (identity holds in all complete rows)")
log_check("pharmacy", "Implausible days supply on paid claims (< 3 or > 90 days)",
          sum(pharmacy$claim_status == "P" & (pharmacy$days_supply < 3 | pharmacy$days_supply > 90)),
          "Replace with the most common days supply for the same product and quantity")
```

**What.** Medical: count duplicates, lower-case and dotted ICD-10 codes, place of service with a lost zero, missing or negative allowed amounts, very large amounts, and obesity-class codes used before their effective date.

**Why.** Diagnosis codes define obesity, diabetes and comorbidities; cost fields define the economic outcomes.

```{r dq-medical}
dx_long <- unlist(medical[, paste0("dx", 1:5)], use.names = FALSE)
dx_long <- dx_long[!is.na(dx_long)]
log_check("medical", "Exact duplicate lines", sum(duplicated(data.table::as.data.table(medical))), "Drop exact duplicates")
log_check("medical", "ICD-10-CM codes in lower case", sum(dx_long != toupper(dx_long)), "Convert to upper case")
log_check("medical", "ICD-10-CM codes with a decimal point (mixed formats)", sum(grepl(".", dx_long, fixed = TRUE)),
          "Remove the decimal point (match on undotted codes)")
log_check("medical", "Place of service with leading zero lost ('2' instead of '02')",
          sum(nchar(medical$place_of_service) == 1, na.rm = TRUE), "Left-pad to 2 characters")
log_check("medical", "Duplicate claim_id + line_num after code normalization",
          sum(duplicated(data.table::as.data.table(medical[, c("claim_id", "line_num")]))) - sum(duplicated(data.table::as.data.table(medical))),
          "Keep one row per claim line")
log_check("medical", "Missing allowed amount", sum(is.na(medical$allowed_amt)),
          "Impute as paid_amt / median(paid/allowed ratio) for the same claim type")
log_check("medical", "Negative allowed amounts (adjustment lines, line_num >= 51)", sum(medical$allowed_amt < 0, na.rm = TRUE),
          "Keep: adjustments net against the claim; claim totals floored at $0")
log_check("medical", "Very large line amounts (> $100,000; inpatient stays)", sum(medical$allowed_amt > 1e5, na.rm = TRUE),
          "Keep (plausible); costs winsorized at the 99th percentile in a sensitivity analysis")
log_check("medical", "Obesity class codes E66.811-E66.813 used before their 2024-10-01 effective date",
          sum(grepl("^E66\\.?81[123]", toupper(medical$dx1)) & medical$service_from_date < as.Date("2024-10-01")),
          "None found; codes accepted from 2024-10-01")
```

**What.** Combine the checks with `bind_rows(dq)` and save `06_output/tables/data_quality_log.csv`.

**Why.** The log is a deliverable and is summarized in the data description document.

```{r dq-save}
dq_log <- bind_rows(dq)
readr::write_csv(dq_log, file.path(dir_tables, "data_quality_log.csv"))
print(dq_log, n = Inf, width = 200)
```

# Step 4. Clean members and enrollment

**What**

- Drop exact duplicate member rows (`distinct()`) and flag IDs that still have more than one row (`demo_conflict = n() > 1`).
- Add census region from the state lookup (`left_join(... state_census_region.csv)`), with "Unknown" when state is missing.

**Why.** One clean row per member is needed before joining; flagged members are excluded at cohort step 4.

```{r clean-members}
members <- readRDS(file.path(dir_derived, "raw_members.rds"))
members_clean <- members |>
  distinct() |>                                   # drop exact duplicate rows
  group_by(member_id) |>
  mutate(demo_conflict = n() > 1) |>              # same ID, different demographics
  slice(1) |>
  ungroup() |>
  left_join(readr::read_csv(file.path(dir_ref, "state_census_region.csv"), show_col_types = FALSE),
            by = "state") |>
  mutate(region = coalesce(census_region, "Unknown")) |>
  select(member_id, birth_year, sex, state, region, demo_conflict)
saveRDS(members_clean, file.path(dir_derived, "members_clean.rds"))
count(members_clean, demo_conflict)
```

**What**

- Keep segments with both medical and pharmacy coverage (`filter(medical_coverage == "Y", rx_coverage == "Y")`).
- Merge overlapping or nearly adjacent segments into continuous spans: `cummax()` tracks the latest end date, and a new span starts only when the next segment begins more than 30 days later (`new_span`).

**Why.** Continuous enrollment (365 days before and after index) guarantees we see all claims in the baseline and follow-up windows, so a missing claim means no care, not missing data.

```{r clean-enrollment}
enrollment <- readRDS(file.path(dir_derived, "raw_enrollment.rds"))

# Covered time = segments with BOTH medical and pharmacy coverage
segments <- enrollment |>
  distinct() |>
  filter(medical_coverage == "Y", rx_coverage == "Y") |>
  arrange(member_id, eligibility_start)
saveRDS(segments, file.path(dir_derived, "enrollment_segments.rds"))  # used for plan type at index

# Collapse overlapping/adjacent segments into continuous spans.
# A new span starts only when a segment begins more than `enroll_gap_allowed` days
# after the latest end date seen so far for that member.
spans <- segments |>
  group_by(member_id) |>
  mutate(run_end   = as.Date(cummax(as.numeric(eligibility_end))),
         new_span  = eligibility_start > lag(run_end) + 1 + enroll_gap_allowed,
         new_span  = coalesce(new_span, TRUE),
         span_id   = cumsum(new_span)) |>
  group_by(member_id, span_id) |>
  summarise(span_start = min(eligibility_start), span_end = max(eligibility_end), .groups = "drop")
saveRDS(spans, file.path(dir_derived, "enrollment_spans.rds"))
message("Segments: ", nrow(segments), " -> continuous spans: ", nrow(spans),
        " for ", n_distinct(spans$member_id), " members")
```

# Step 5. Clean pharmacy and medical claims

**What**

- Convert every NDC to 11 digits with `normalize_ndc()` (remove hyphens, pad the labeler, restore lost leading zeros).
- Remove reversed claims and the paid claims they reverse (`!rx_claim_id %in% reversed_ids`), and keep one row per claim (`distinct(rx_claim_id, ...)`).
- Fill missing allowed amounts as paid + out-of-pocket (`coalesce()`), and add drug names from the lookup (`left_join(ndc_lookup ...)`).
- Replace implausible days supply with the most common value for the same NDC and quantity (`ds_mode`, `days_supply_fixed`).
- Check that every NDC matched (`stopifnot()`).

**Why.** Without these fixes some Zepbound and Wegovy fills would be missed, reversed fills would count as treatment, and a "999-day" supply would make a patient look persistent for years.

```{r clean-pharmacy}
pharmacy   <- readRDS(file.path(dir_derived, "raw_pharmacy.rds"))
ndc_lookup <- readr::read_csv(file.path(dir_ref, "ndc_product_lookup.csv"),
                              col_types = cols(.default = col_character()))

# NDCs arrive in three formats; convert all to the 11-digit (5-4-2) form used by the lookup
normalize_ndc <- function(x) {
  digits <- gsub("-", "", x)
  out <- ifelse(grepl("-", x) & nchar(x) == 12, paste0("0", digits), digits)  # 4-4-2 -> 5-4-2
  stringr::str_pad(out, 11, side = "left", pad = "0")                         # lost leading zeros
}

reversed_ids <- pharmacy$rx_claim_id[pharmacy$claim_status == "R"]

pharmacy_clean <- pharmacy |>
  mutate(ndc11 = normalize_ndc(ndc)) |>
  filter(claim_status == "P", !rx_claim_id %in% reversed_ids) |>  # drop reversals and what they reverse
  distinct(rx_claim_id, .keep_all = TRUE) |>                       # one row per paid claim
  mutate(allowed_amt = coalesce(allowed_amt, paid_amt + member_oop)) |>
  left_join(ndc_lookup |> select(ndc11, product_name, generic_name, strength, therapeutic_class),
            by = "ndc11")

# Implausible days supply (< 3 or > 90): use the most common value for the same NDC and quantity
mode_val <- function(x) as.integer(names(which.max(table(x))))
ds_mode <- pharmacy_clean |>
  filter(between(days_supply, 3, 90)) |>
  group_by(ndc11, quantity) |>
  summarise(ds_typical = mode_val(days_supply), .groups = "drop")
ds_mode_ndc <- pharmacy_clean |>
  filter(between(days_supply, 3, 90)) |>
  group_by(ndc11) |>
  summarise(ds_typical_ndc = mode_val(days_supply), .groups = "drop")

pharmacy_clean <- pharmacy_clean |>
  left_join(ds_mode, by = c("ndc11", "quantity")) |>
  left_join(ds_mode_ndc, by = "ndc11") |>
  mutate(days_supply_fixed = !between(days_supply, 3, 90),
         days_supply = if_else(days_supply_fixed, coalesce(ds_typical, ds_typical_ndc), days_supply)) |>
  select(-ds_typical, -ds_typical_ndc)

stopifnot(all(!is.na(pharmacy_clean$product_name)))   # every NDC matched the lookup
saveRDS(pharmacy_clean, file.path(dir_derived, "pharmacy_clean.rds"))
message("Pharmacy: ", nrow(pharmacy), " raw rows -> ", nrow(pharmacy_clean), " clean paid claims; ",
        sum(pharmacy_clean$days_supply_fixed), " days-supply values corrected")
```

**What**

- Standardize diagnosis codes (`clean_dx()`: upper case, no dot) and pad place of service to two characters (`str_pad()`).
- Remove duplicate lines (`distinct()`, then one row per `claim_id` + `line_num`).
- Back-calculate missing allowed amounts from paid amounts using the median paid/allowed ratio by claim type (`paid_ratio`).

**Why.** Consistent codes make the code lists work, and duplicate lines would double-count utilization and costs.

```{r clean-medical}
medical <- readRDS(file.path(dir_derived, "raw_medical.rds"))

clean_dx <- function(x) gsub(".", "", toupper(trimws(x)), fixed = TRUE)

medical_clean <- medical |>
  mutate(across(dx1:dx5, clean_dx),
         place_of_service = stringr::str_pad(place_of_service, 2, side = "left", pad = "0"),
         proc_code = toupper(proc_code)) |>
  distinct() |>                                     # exact duplicates (incl. after normalization)
  distinct(claim_id, line_num, .keep_all = TRUE)    # one row per claim line

# Missing allowed amounts: back-calculate from paid using the typical paid/allowed ratio
paid_ratio <- medical_clean |>
  filter(allowed_amt > 0) |>
  group_by(claim_type) |>
  summarise(ratio = median(paid_amt / allowed_amt), .groups = "drop")
medical_clean <- medical_clean |>
  left_join(paid_ratio, by = "claim_type") |>
  mutate(allowed_imputed = is.na(allowed_amt),
         allowed_amt = if_else(allowed_imputed, round(paid_amt / ratio, 2), allowed_amt)) |>
  select(-ratio)

saveRDS(medical_clean, file.path(dir_derived, "medical_clean.rds"))
message("Medical: ", nrow(medical), " raw lines -> ", nrow(medical_clean), " clean lines; ",
        sum(medical_clean$allowed_imputed), " allowed amounts imputed")
```

# Step 6. Build the cohort (with a flow diagram)

**What**

- Load the clean tables and build a long diagnosis table, one row per member, date and code (`pivot_longer(dx1:dx5)`).

**Why.** Searching one column is simpler and faster than searching five.

```{r cohort-load}
members  <- readRDS(file.path(dir_derived, "members_clean.rds"))
spans    <- readRDS(file.path(dir_derived, "enrollment_spans.rds"))
segments <- readRDS(file.path(dir_derived, "enrollment_segments.rds"))
rx       <- readRDS(file.path(dir_derived, "pharmacy_clean.rds"))
med      <- readRDS(file.path(dir_derived, "medical_clean.rds"))

# Long diagnosis table: one row per member, date and diagnosis code (any position dx1-dx5)
dx_long <- med |>
  select(member_id, service_from_date, dx1:dx5) |>
  tidyr::pivot_longer(dx1:dx5, values_to = "dx", values_drop_na = TRUE) |>
  distinct(member_id, date = service_from_date, dx)
saveRDS(dx_long, file.path(dir_derived, "dx_long.rds"))
```

**What**

- Define the code lists: study products, all GLP-1 RA products, obesity (`dx_obesity`), diabetes (`dx_diabetes`), contraindications (`dx_contraind`) and non-metformin antidiabetic drug classes.
- Write a helper, `has_dx()`, that returns TRUE if a member has a matching code between two dates.

**Why.** Code lists written in one place are easy to review against the SAP and reuse.

```{r cohort-codes}
# Code lists (ICD-10-CM without decimal point; regular expressions on the start of the code)
study_products  <- c(ZEPBOUND = "Tirzepatide", WEGOVY = "Semaglutide 2.4 mg")
glp1_products   <- c("ZEPBOUND", "WEGOVY", "MOUNJARO", "OZEMPIC", "RYBELSUS", "SAXENDA")
dx_obesity      <- "^E66(0|1|2|8|9)|^Z68(3|4)"   # E66.x except E66.3 overweight; BMI >= 30
dx_diabetes     <- "^E1[01]"                     # type 1 (E10) or type 2 (E11) diabetes
dx_contraind    <- "^C73|^E3122"                 # thyroid cancer / MEN 2 (boxed warning)
antidiabetic_classes <- c("SGLT2 INHIBITOR", "SULFONYLUREA", "INSULIN, LONG-ACTING", "DPP-4 INHIBITOR")

# helper: does a member have a code matching `pattern` between two dates?
has_dx <- function(ids, from, to, pattern) {
  hits <- dx_long |> filter(grepl(pattern, dx)) |> select(member_id, date)
  tibble(member_id = ids, from = from, to = to) |>
    left_join(hits, by = "member_id", relationship = "many-to-many") |>
    group_by(member_id) |>
    summarise(flag = any(!is.na(date) & date >= from & date <= to), .groups = "drop") |>
    pull(flag, name = member_id) |> (\(x) x[ids])()
}
```

**What**

- Find each member's first Zepbound or Wegovy fill in January 2024 to June 2025: this is the index date and sets the treatment group (`summarise(index_date = first(fill_date))`).
- Record the counts at each step with `add_step()`.

**Why.** In the target trial emulation, the first fill is "time zero": eligibility, treatment assignment and the start of follow-up all happen on the same day, which avoids immortal time bias.

```{r cohort-index}
attrition <- list()
add_step <- function(step, data) {
  attrition[[length(attrition) + 1]] <<- tibble(
    step = step, n = nrow(data),
    n_tirzepatide = sum(data$drug == "Tirzepatide"),
    n_semaglutide = sum(data$drug == "Semaglutide 2.4 mg"))
  data
}

study_fills <- rx |> filter(product_name %in% names(study_products))

# Index = first Zepbound or Wegovy fill within the identification window
index <- study_fills |>
  filter(fill_date >= index_start, fill_date <= index_end) |>
  arrange(member_id, fill_date) |>
  group_by(member_id) |>
  summarise(index_date = first(fill_date),
            n_products_on_index = n_distinct(product_name[fill_date == first(fill_date)]),
            index_product = first(product_name), .groups = "drop") |>
  mutate(drug = factor(study_products[index_product], levels = c("Semaglutide 2.4 mg", "Tirzepatide")))

cohort <- add_step("1. First Zepbound or Wegovy fill, Jan 2024 - Jun 2025 (index date)", index)
```

**What.** Apply the criteria in SAP order, recording counts after each `filter()` / `semi_join()` / `anti_join()`:

1. one study drug on the index date;
2. age 18-64;
3. known sex and no conflicting records;
4. continuous enrollment (`span_start <= index_date - 365`, `span_end >= index_date + 364`);
5. no GLP-1 RA fill in the 365-day baseline (new user);
6. obesity diagnosis;
7. no diabetes (no E10/E11 code and no non-metformin antidiabetic fill);
8. no thyroid cancer / MEN 2 code.

**Why.** Each criterion puts into practice an element of the PECO question and the target trial. The new-user washout makes the two groups comparable at the start of treatment.

```{r cohort-criteria}
cohort <- cohort |>
  filter(n_products_on_index == 1) |>
  add_step("2. Only one study drug filled on the index date", data = _)

cohort <- cohort |>
  left_join(members, by = "member_id") |>
  mutate(age = lubridate::year(index_date) - birth_year) |>
  filter(!is.na(age), age >= 18, age <= 64) |>
  add_step("3. Aged 18-64 at index (birth year recorded)", data = _)

cohort <- cohort |>
  filter(!demo_conflict, sex %in% c("F", "M")) |>
  add_step("4. Known sex and no conflicting member records", data = _)

cont_enrolled <- spans |>
  inner_join(cohort |> select(member_id, index_date), by = "member_id") |>
  filter(span_start <= index_date - baseline_days, span_end >= index_date + followup_days - 1) |>
  distinct(member_id)
cohort <- cohort |>
  semi_join(cont_enrolled, by = "member_id") |>
  add_step("5. Continuous medical + pharmacy enrollment 365 days before and after index", data = _)

prior_glp1 <- rx |>
  filter(product_name %in% glp1_products) |>
  inner_join(cohort |> select(member_id, index_date), by = "member_id") |>
  filter(fill_date >= index_date - baseline_days, fill_date < index_date) |>
  distinct(member_id)
cohort <- cohort |>
  anti_join(prior_glp1, by = "member_id") |>
  add_step("6. New user: no GLP-1 RA fill (any product) in the 365-day baseline", data = _)

cohort <- cohort |>
  filter(has_dx(member_id, index_date - baseline_days, index_date, dx_obesity)) |>
  add_step("7. Obesity diagnosis (E66 excl. E66.3, or BMI Z68.30-Z68.45) in baseline or on index", data = _)

baseline_antidiabetic <- rx |>
  filter(therapeutic_class %in% antidiabetic_classes) |>
  inner_join(cohort |> select(member_id, index_date), by = "member_id") |>
  filter(fill_date >= index_date - baseline_days, fill_date <= index_date) |>
  distinct(member_id)
cohort <- cohort |>
  filter(!has_dx(member_id, index_date - baseline_days, index_date, dx_diabetes)) |>
  anti_join(baseline_antidiabetic, by = "member_id") |>
  add_step("8. No diabetes (no E10/E11 diagnosis, no non-metformin antidiabetic drug) in baseline", data = _)

cohort <- cohort |>
  filter(!has_dx(member_id, index_date - baseline_days, index_date, dx_contraind)) |>
  add_step("9. No medullary thyroid cancer / MEN 2 risk code (C73, E31.22) in baseline", data = _)
```

**What.** Add plan type and employer group from the enrollment segment covering the index date (`eligibility_start <= index_date`, `eligibility_end >= index_date`), add the index quarter, and save `cohort.rds`.

**Why.** Plan type is a confounder (benefit design and cost sharing). Index quarter captures calendar-time changes in supply and coverage.

```{r cohort-plan}
# Plan type and employer group from the enrollment segment covering the index date
plan_at_index <- segments |>
  inner_join(cohort |> select(member_id, index_date), by = "member_id") |>
  filter(eligibility_start <= index_date, eligibility_end >= index_date) |>
  arrange(member_id, desc(eligibility_start)) |>
  distinct(member_id, .keep_all = TRUE) |>
  select(member_id, plan_type, group_id)

cohort <- cohort |>
  left_join(plan_at_index, by = "member_id") |>
  mutate(index_qtr = paste0(lubridate::year(index_date), "Q", lubridate::quarter(index_date)),
         treat = as.integer(drug == "Tirzepatide")) |>
  select(member_id, index_date, index_qtr, drug, treat, age, sex, state, region, plan_type, group_id)
stopifnot(!anyNA(cohort$plan_type), !anyDuplicated(cohort$member_id))
saveRDS(cohort, file.path(dir_derived, "cohort.rds"))
```

**What.** Save the attrition table and draw the flow diagram (Figure 1) with ggplot2.

**Why.** The flow diagram is required by RECORD-PE and shows readers how the study population was reached.

```{r cohort-attrition, fig.height = 8}
attrition_tbl <- bind_rows(attrition) |>
  mutate(excluded = lag(n) - n)
readr::write_csv(attrition_tbl, file.path(dir_tables, "attrition.csv"))
print(attrition_tbl, width = 200)

# Flow diagram (Figure 1)
flow <- attrition_tbl |>
  mutate(y = rev(seq_len(n())),
         label = paste0(step, "\nN = ", scales::comma(n),
                        "  (tirzepatide ", scales::comma(n_tirzepatide),
                        "; semaglutide ", scales::comma(n_semaglutide), ")"),
         excl  = ifelse(is.na(excluded), NA, paste0("Excluded: ", scales::comma(excluded))))
p_flow <- ggplot(flow, aes(x = 0, y = y)) +
  geom_segment(data = flow[-1, ], aes(x = 0, xend = 0, y = y + 0.62, yend = y + 0.38),
               arrow = arrow(length = unit(2, "mm"))) +
  geom_label(aes(label = label), size = 2.7, linewidth = 0.3, fill = "white", hjust = 0.5) +
  geom_text(aes(x = 0.9, y = y + 0.5, label = excl), size = 2.6, hjust = 0, colour = "grey30", na.rm = TRUE) +
  scale_x_continuous(limits = c(-1.1, 1.3)) +
  theme_void() +
  labs(title = "Figure 1. Cohort selection (synthetic commercial claims, 2023-2026)")
ggsave(file.path(dir_figures, "fig1_cohort_flow.png"), p_flow, width = 9, height = 8, dpi = 200, bg = "white")
```

# Step 7. Baseline covariates

**What.** Load the cohort and claims, and compute each claim's day relative to the index date (`day = fill_date - index_date`).

**Why.** Working in "days since index" makes every time window a simple filter (e.g., baseline = `day >= -365 & day <= -1`).

```{r cov-load}
cohort  <- readRDS(file.path(dir_derived, "cohort.rds"))
rx      <- readRDS(file.path(dir_derived, "pharmacy_clean.rds"))
med     <- readRDS(file.path(dir_derived, "medical_clean.rds"))
dx_long <- readRDS(file.path(dir_derived, "dx_long.rds"))

ids <- cohort |> select(member_id, index_date)
# Keep only cohort members' claims, with each claim's day relative to index (0 = index date)
rx_c  <- rx  |> inner_join(ids, by = "member_id") |> mutate(day = as.integer(fill_date - index_date))
med_c <- med |> inner_join(ids, by = "member_id") |> mutate(day = as.integer(service_from_date - index_date))
dx_c  <- dx_long |> inner_join(ids, by = "member_id") |> mutate(day = as.integer(date - index_date))
```

**What**

- Define encounters once, one per member per day: inpatient admissions (`claim_type == "INST"` with an admission date), ED visits (place of service 23, revenue code 0450 or CPT 99281-99285), outpatient E/M visits, and obesity-related outpatient visits (E66/Z68 on the claim).
- Sum medical costs by claim and floor each claim at $0 (`max(0, sum(allowed_amt))`).

**Why.** The same definitions are used for baseline covariates (here) and follow-up outcomes (Step 8), so the two are measured consistently.

```{r cov-encounters}
# Encounter types used for baseline AND follow-up utilization (one visit per member per day)
em_outpatient <- c(sprintf("%d", c(99202:99205, 99211:99215, 99381:99397)), "G0438")
encounters <- bind_rows(
  med_c |> filter(claim_type == "INST", !is.na(admit_date)) |>
    distinct(member_id, day = as.integer(admit_date - index_date)) |> mutate(type = "inpatient"),
  med_c |> filter((claim_type == "INST" & place_of_service == "23") | revenue_code %in% "0450" |
                    proc_code %in% as.character(99281:99285)) |>
    distinct(member_id, day) |> mutate(type = "ed"),
  med_c |> filter(claim_type == "PROF", proc_code %in% em_outpatient,
                  place_of_service %in% c("02", "10", "11", "19", "22")) |>
    distinct(member_id, day) |> mutate(type = "outpatient"),
  med_c |> filter(claim_type == "PROF", proc_code %in% em_outpatient,
                  place_of_service %in% c("02", "10", "11", "19", "22"),
                  if_any(dx1:dx5, ~ grepl("^E66|^Z68", .x))) |>
    distinct(member_id, day) |> mutate(type = "outpatient_obesity")
)
saveRDS(encounters, file.path(dir_derived, "encounters.rds"))

# Claim-level medical costs: adjustment lines net against their claim; floor at $0
med_claims <- med_c |>
  group_by(member_id, claim_id) |>
  summarise(day = min(day), allowed = max(0, sum(allowed_amt)), .groups = "drop")
saveRDS(med_claims, file.path(dir_derived, "med_claims.rds"))
```

**What**

- Flag comorbidities from codes in [index - 365, index] with `dx_flag()` (e.g., hypertension `^I1[0-6]`, sleep apnea `^G4733`).
- Assign obesity class from the most recent BMI or class code (`case_when()`; Class 3 = BMI >= 40, E66.01, E66.2 or E66.813).

**Why.** These are the confounders in the DAG: they influence which drug is prescribed and the outcomes.

```{r cov-diagnoses}
# Diagnosis flags: any code in [index - 365, index]
dx_base <- dx_c |> filter(day >= -baseline_days, day <= 0)
dx_flag <- function(pattern) ids$member_id %in% dx_base$member_id[grepl(pattern, dx_base$dx)]
comorb <- tibble(
  member_id     = ids$member_id,
  hypertension  = dx_flag("^I1[0-6]"),
  dyslipidemia  = dx_flag("^E78"),
  sleep_apnea   = dx_flag("^G4733"),
  prediabetes   = dx_flag("^R730"),
  masld         = dx_flag("^K760|^K7581"),
  osteoarthritis = dx_flag("^M1[5-9]"),
  depression    = dx_flag("^F3[23]"),
  anxiety       = dx_flag("^F41"),
  gerd          = dx_flag("^K21"),
  ascvd         = dx_flag("^I2[0-5]|^I63|^I73|^Z8673"),
  heart_failure = dx_flag("^I50"),
  ckd           = dx_flag("^N18"),
  pcos          = dx_flag("^E282"),
  hypothyroid   = dx_flag("^E03"),
  gallbladder   = dx_flag("^K80"),
  pancreatitis  = dx_flag("^K8[56]")
)

# Obesity class from the most recent BMI or class-specific code in baseline
obesity_class <- dx_base |>
  mutate(cls = case_when(
    grepl("^Z68(4[1-5])|^E66813|^E6601|^E662", dx) ~ 3L,
    grepl("^Z683[5-9]|^E66812", dx)                ~ 2L,
    grepl("^Z683[0-4]|^E66811", dx)                ~ 1L)) |>
  filter(!is.na(cls)) |>
  group_by(member_id) |>
  filter(day == max(day)) |>
  summarise(cls = max(cls), .groups = "drop")
```

**What.** Baseline utilization and costs: counts of each encounter type (`count(member_id, type)`), baseline medical and pharmacy costs, number of distinct generic drugs, endocrinologist visit, and prior bariatric surgery.

**Why.** Prior healthcare use is the best available proxy for overall health and care-seeking, which affect both drug choice and future costs.

```{r cov-utilization}
# Baseline utilization and costs: [index - 365, index - 1]
base_enc <- encounters |>
  filter(day >= -baseline_days, day <= -1) |>
  count(member_id, type) |>
  tidyr::pivot_wider(names_from = type, values_from = n, values_fill = 0)
base_med_cost <- med_claims |> filter(day >= -baseline_days, day <= -1) |>
  group_by(member_id) |> summarise(base_med_cost = sum(allowed))
base_rx_cost  <- rx_c |> filter(day >= -baseline_days, day <= -1) |>
  group_by(member_id) |> summarise(base_rx_cost = sum(allowed_amt), n_generics = n_distinct(generic_name))
endo_visit    <- med_c |> filter(day >= -baseline_days, day <= -1, prov_specialty == "END") |> distinct(member_id)
bariatric     <- union(
  dx_base$member_id[grepl("^Z9884", dx_base$dx)],
  med_c$member_id[med_c$proc_code %in% c("43644", "43775") & med_c$day < 0 & med_c$day >= -baseline_days])
```

**What.** Baseline medication flags by therapeutic class with `rx_flag()` (other anti-obesity drugs, metformin, antihypertensives, statins, antidepressants).

**Why.** Medication use marks comorbidity severity and treatment history.

```{r cov-medications}
rx_base <- rx_c |> filter(day >= -baseline_days, day <= -1)
rx_flag <- function(classes) ids$member_id %in% rx_base$member_id[rx_base$therapeutic_class %in% classes]
meds <- tibble(
  member_id         = ids$member_id,
  prior_aom         = rx_flag(c("ANOREXIGENIC AGENT", "ANTI-OBESITY AGENT")),
  metformin         = rx_flag("BIGUANIDE"),
  antihypertensive  = rx_flag(c("ACE INHIBITOR", "ANGIOTENSIN II RECEPTOR BLOCKER", "CALCIUM CHANNEL BLOCKER",
                                "BETA BLOCKER", "LOOP DIURETIC", "ALDOSTERONE ANTAGONIST")),
  statin            = rx_flag("HMG-COA REDUCTASE INHIBITOR"),
  antidepressant    = rx_flag(c("SSRI", "ANTIDEPRESSANT, OTHER"))
)
```

**What.** Join everything into one row per patient; create factors (obesity class, age group, CDHP vs non-CDHP), any baseline inpatient/ED use, and `log_base_cost = log1p(base_total_cost)`; save `covariates.rds`.

**Why.** The log of baseline cost reduces the influence of a few very expensive patients in the propensity score model.

```{r cov-assemble}
covariates <- cohort |>
  left_join(comorb, by = "member_id") |>
  left_join(meds, by = "member_id") |>
  left_join(obesity_class, by = "member_id") |>
  left_join(base_enc, by = "member_id") |>
  left_join(base_med_cost, by = "member_id") |>
  left_join(base_rx_cost, by = "member_id") |>
  mutate(
    across(c(inpatient, ed, outpatient, outpatient_obesity, base_med_cost, base_rx_cost, n_generics),
           ~ coalesce(.x, 0)),
    obesity_class   = factor(coalesce(c("Class 1", "Class 2", "Class 3")[cls], "Unknown"),
                             levels = c("Class 1", "Class 2", "Class 3", "Unknown")),
    age_group       = factor(ifelse(age < 45, "18-44", "45-64")),
    sex             = factor(sex, levels = c("F", "M"), labels = c("Female", "Male")),
    region          = factor(region),
    plan_type       = factor(plan_type, levels = c("PPO", "HMO", "POS", "EPO", "CDHP")),
    cdhp            = factor(ifelse(plan_type == "CDHP", "CDHP", "Non-CDHP"), levels = c("Non-CDHP", "CDHP")),
    index_qtr       = factor(index_qtr),
    endo_visit      = member_id %in% endo_visit$member_id,
    bariatric_hx    = member_id %in% bariatric,
    base_any_ip     = inpatient > 0,
    base_any_ed     = ed > 0,
    base_op_visits  = outpatient,
    base_total_cost = base_med_cost + base_rx_cost,
    log_base_cost   = log1p(base_total_cost)
  ) |>
  select(-cls, -inpatient, -ed, -outpatient, -outpatient_obesity)
stopifnot(nrow(covariates) == nrow(cohort))
saveRDS(covariates, file.path(dir_derived, "covariates.rds"))
glimpse(covariates)
```

# Step 8. Outcomes over the 365-day follow-up

**What.** Load data; keep follow-up fills (days 0-364) of either molecule, labelled by molecule (Zepbound/Mounjaro = tirzepatide; Wegovy/Ozempic/Rybelsus = semaglutide).

**Why.** Persistence is measured on the index molecule, and fills of the other molecule mark switching.

```{r out-load}
covariates <- readRDS(file.path(dir_derived, "covariates.rds"))
rx         <- readRDS(file.path(dir_derived, "pharmacy_clean.rds"))
encounters <- readRDS(file.path(dir_derived, "encounters.rds"))
med_claims <- readRDS(file.path(dir_derived, "med_claims.rds"))
med        <- readRDS(file.path(dir_derived, "medical_clean.rds"))

ids <- covariates |> select(member_id, index_date, drug)
fu_last <- followup_days - 1   # follow-up runs from day 0 (index) to day 364

# Follow-up fills of tirzepatide or semaglutide (any brand) with the day relative to index
molecule <- c(ZEPBOUND = "Tirzepatide", MOUNJARO = "Tirzepatide",
              WEGOVY = "Semaglutide 2.4 mg", OZEMPIC = "Semaglutide 2.4 mg", RYBELSUS = "Semaglutide 2.4 mg")
fu_fills <- rx |>
  filter(product_name %in% names(molecule)) |>
  inner_join(ids, by = "member_id") |>
  mutate(day = as.integer(fill_date - index_date), fill_molecule = molecule[product_name]) |>
  filter(day >= 0, day <= fu_last) |>
  arrange(member_id, day)
```

**What.** Write the persistence function `persistence(day, ds, gap)`:

- `start`/`end` lay the fills end to end, carrying early refills forward;
- `gaps_before` and `tail_gap` measure days without supply;
- the first gap longer than `gap` days is a discontinuation (`disc_day`);
- `pdc` = days covered / 365.

**Why.** This is the standard claims definition of persistence (no refill gap > 60 days) and adherence (proportion of days covered), as used by Marshall 2026. Writing it as a function lets us change the gap for the sensitivity analyses.

```{r out-persistence-function}
# Persistence for one patient's fills of a drug.
# - Early refills are carried over: a fill's supply starts the day after the previous supply ends.
# - Discontinuation = first gap of more than `gap` days without supply (including the gap
#   between the last supply and the end of follow-up). Event day = first uncovered day.
# - PDC = days covered in the 365-day follow-up / 365.
persistence <- function(day, ds, gap = 60, last = 364) {
  n <- length(day); start <- end <- integer(n)
  for (k in seq_len(n)) {
    start[k] <- if (k == 1) day[k] else max(day[k], end[k - 1] + 1L)
    end[k]   <- start[k] + ds[k] - 1L
  }
  gaps_before <- c(0L, start[-1] - end[-n] - 1L)          # gap before each refill
  tail_gap    <- last - end[n]                            # gap after the last supply
  k_gap <- which(gaps_before > gap)[1]
  disc_day <- if (!is.na(k_gap)) end[k_gap - 1] + 1L else if (tail_gap > gap) end[n] + 1L else NA_integer_
  covered  <- sum(pmax(0L, pmin(end, last) - pmin(start, last + 1L) + 1L))
  tibble(persistent = is.na(disc_day),
         time_to_disc = if (is.na(disc_day)) as.integer(last + 1) else min(disc_day, last + 1L),
         pdc = covered / (last + 1))
}
```

**What.** Apply the function to each patient (`group_by(member_id) |> reframe(persistence(...))`) with gaps of 60 (primary), 30 and 90 days, and on either molecule (`p_class`). Flag switching to the other molecule (`switching`).

**Why.** The primary outcome and sensitivity analyses S1-S3 come from the same function, so only the definition changes.

```{r out-persistence}
index_fills <- fu_fills |> filter(fill_molecule == drug)   # the index molecule only

persist_all <- function(fills, gap) {
  fills |>
    group_by(member_id) |>
    reframe(persistence(day, days_supply, gap = gap, last = fu_last))
}
p60 <- persist_all(index_fills, 60)
p30 <- persist_all(index_fills, 30) |> rename_with(~ paste0(.x, "_g30"), -member_id)
p90 <- persist_all(index_fills, 90) |> rename_with(~ paste0(.x, "_g90"), -member_id)
# Sensitivity: persistence on EITHER molecule (switching between them is not discontinuation)
p_class <- persist_all(fu_fills, 60) |>
  select(member_id, persistent_class = persistent, time_to_disc_class = time_to_disc)

switching <- fu_fills |>
  filter(fill_molecule != drug) |>
  group_by(member_id) |>
  summarise(switched = TRUE, switch_day = min(day))
```

**What.** Count follow-up encounters by type, and sum 12-month medical, pharmacy, study-drug and out-of-pocket costs. Flag the negative control outcome (colonoscopy 45378 or screening mammogram 77067).

**Why.** These are the Aim 3 and 4 outcomes. The negative control outcome checks for residual confounding by healthcare-seeking behaviour.

```{r out-utilization-costs}
fu_enc <- encounters |>
  filter(day >= 0, day <= fu_last) |>
  count(member_id, type) |>
  tidyr::pivot_wider(names_from = type, values_from = n, values_fill = 0, names_prefix = "n_")

fu_med_cost <- med_claims |> filter(day >= 0, day <= fu_last) |>
  group_by(member_id) |> summarise(cost_medical = sum(allowed))
fu_rx_cost  <- rx |> inner_join(ids, by = "member_id") |>
  mutate(day = as.integer(fill_date - index_date)) |>
  filter(day >= 0, day <= fu_last) |>
  group_by(member_id) |>
  summarise(cost_pharmacy = sum(allowed_amt),
            cost_study_drug = sum(allowed_amt[product_name %in% c("ZEPBOUND", "WEGOVY")]),
            oop_study_drug  = sum(member_oop[product_name %in% c("ZEPBOUND", "WEGOVY")]))

# Negative control outcome: routine cancer screening (colonoscopy 45378, screening mammogram 77067)
neg_control <- med |> filter(proc_code %in% c("45378", "77067")) |>
  inner_join(ids, by = "member_id") |>
  mutate(day = as.integer(service_from_date - index_date)) |>
  filter(day >= 0, day <= fu_last) |>
  distinct(member_id) |> mutate(neg_control = TRUE)
```

**What.** Join all outcomes, fill zeros for patients with no encounters (`coalesce(.x, 0)`), create binary outcomes (`pdc80`, `any_ip`, `any_ed`) and total cost; save `analytic.rds`; print a crude (unweighted) summary by drug.

**Why.** One analytic dataset with one row per patient feeds every model. The crude summary is a first look before adjustment.

```{r out-assemble}
outcomes <- ids |>
  select(member_id) |>
  left_join(p60, by = "member_id") |> left_join(p30, by = "member_id") |> left_join(p90, by = "member_id") |>
  left_join(p_class, by = "member_id") |> left_join(switching, by = "member_id") |>
  left_join(fu_enc, by = "member_id") |> left_join(fu_med_cost, by = "member_id") |>
  left_join(fu_rx_cost, by = "member_id") |> left_join(neg_control, by = "member_id") |>
  mutate(
    across(c(switched, neg_control), ~ coalesce(.x, FALSE)),
    across(c(n_inpatient, n_ed, n_outpatient, n_outpatient_obesity, cost_medical), ~ coalesce(.x, 0)),
    discontinued = !persistent,
    pdc80        = pdc >= 0.80,
    any_ip       = n_inpatient > 0,
    any_ed       = n_ed > 0,
    cost_total   = cost_medical + cost_pharmacy
  )
stopifnot(nrow(outcomes) == nrow(ids), !anyNA(outcomes$persistent), all(outcomes$cost_pharmacy > 0))
saveRDS(outcomes, file.path(dir_derived, "outcomes.rds"))

analytic <- covariates |> left_join(outcomes, by = "member_id")
saveRDS(analytic, file.path(dir_derived, "analytic.rds"))

# Crude (unweighted) outcome summary by drug
analytic |>
  group_by(drug) |>
  summarise(n = n(), persistent_12m = mean(persistent), mean_pdc = mean(pdc), pdc80 = mean(pdc80),
            switched = mean(switched), any_ip = mean(any_ip), any_ed = mean(any_ed),
            op_visits = mean(n_outpatient), cost_medical = mean(cost_medical),
            cost_pharmacy = mean(cost_pharmacy), cost_total = mean(cost_total)) |>
  print(width = 200)
```

# Step 9. Table 1, propensity score and weighting

**What.** Load the analytic file and write the propensity score formula from the DAG adjustment set (`covs`, `reformulate(covs, response = "treat")`).

**Why.** The PS model includes the confounders and nothing measured after the index date.

```{r ps-formula}
analytic <- readRDS(file.path(dir_derived, "analytic.rds"))

# Adjustment set from the DAG (SAP section 7)
covs <- c("age", "sex", "region", "plan_type", "index_qtr", "obesity_class",
          "hypertension", "dyslipidemia", "sleep_apnea", "prediabetes", "masld", "osteoarthritis",
          "depression", "anxiety", "gerd", "ascvd", "heart_failure", "ckd", "pcos", "hypothyroid",
          "gallbladder", "pancreatitis", "prior_aom", "metformin", "antihypertensive", "statin",
          "antidepressant", "n_generics", "bariatric_hx", "endo_visit", "base_any_ip", "base_any_ed",
          "base_op_visits", "log_base_cost")
ps_formula <- reformulate(covs, response = "treat")
```

**What.** Fit a logistic propensity score model and compute stabilized ATE weights with `WeightIt::weightit(..., method = "glm", estimand = "ATE", stabilize = TRUE)`; save the weights.

**Why.** Inverse probability of treatment weighting creates a pseudo-population in which measured confounders are balanced between the drugs, as in a randomized trial. Stabilization keeps the weights near 1.

```{r ps-weights}
# Propensity score (logistic regression) and stabilized ATE weights
w_ate <- WeightIt::weightit(ps_formula, data = analytic, method = "glm",
                            estimand = "ATE", stabilize = TRUE)
summary(w_ate)
analytic$ps <- w_ate$ps
analytic$w  <- w_ate$weights
saveRDS(analytic, file.path(dir_derived, "analytic_weighted.rds"))
saveRDS(list(covs = covs, ps_formula = ps_formula), file.path(dir_derived, "ps_spec.rds"))
```

**What.** Check balance with `cobalt::bal.tab()`: standardized mean differences (SMD) and variance ratios before and after weighting; report the largest SMD and the effective sample size.

**Why.** Weighting only works if balance is achieved; the SAP target is |SMD| < 0.10 for every covariate.

```{r ps-balance}
bal <- cobalt::bal.tab(w_ate, un = TRUE, binary = "std", continuous = "std",
                       thresholds = c(m = 0.1, v = 2), stats = c("m", "v"))
print(bal)
bal_df <- bal$Balance |>
  tibble::rownames_to_column("covariate") |>
  select(covariate, smd_unweighted = Diff.Un, smd_weighted = Diff.Adj,
         var_ratio_unweighted = V.Ratio.Un, var_ratio_weighted = V.Ratio.Adj)
readr::write_csv(bal_df, file.path(dir_tables, "balance_smd.csv"))
max_smd <- max(abs(bal_df$smd_weighted), na.rm = TRUE)
message("Largest absolute weighted SMD: ", round(max_smd, 3),
        if (max_smd < 0.1) "  -> balance achieved (all < 0.10)" else "  -> IMBALANCE: revise PS model")
ess <- cobalt::bal.tab(w_ate)$Observations
print(ess)
```

**What.** Build Table 1: unweighted and weighted means or percentages by group, with SMDs (`make_rows()`); save to CSV.

**Why.** Table 1 describes the study population and shows the reader that the weighted groups are comparable.

```{r table1}
# Table 1: unweighted and IPTW-weighted characteristics with SMDs
wmean <- function(x, w) sum(x * w) / sum(w)
wsd   <- function(x, w) sqrt(sum(w * (x - wmean(x, w))^2) / sum(w))
t1_vars <- c(covs, "base_med_cost", "base_rx_cost", "base_total_cost")
make_rows <- function(v) {
  x <- analytic[[v]]
  if (is.factor(x) || is.character(x)) {
    lv <- levels(factor(x))
    purrr::map_dfr(lv, function(l) {
      ind <- as.numeric(x == l)
      tibble(variable = v, level = l, type = "pct",
             tirz = mean(ind[analytic$treat == 1]) * 100, sema = mean(ind[analytic$treat == 0]) * 100,
             tirz_w = wmean(ind[analytic$treat == 1], analytic$w[analytic$treat == 1]) * 100,
             sema_w = wmean(ind[analytic$treat == 0], analytic$w[analytic$treat == 0]) * 100,
             tirz_sd = NA, sema_sd = NA)
    })
  } else {
    x <- as.numeric(x); t1 <- analytic$treat == 1; t0 <- !t1
    tibble(variable = v, level = "", type = if (is.logical(analytic[[v]])) "pct" else "mean",
           tirz = mean(x[t1]) * if (is.logical(analytic[[v]])) 100 else 1,
           sema = mean(x[t0]) * if (is.logical(analytic[[v]])) 100 else 1,
           tirz_w = wmean(x[t1], analytic$w[t1]) * if (is.logical(analytic[[v]])) 100 else 1,
           sema_w = wmean(x[t0], analytic$w[t0]) * if (is.logical(analytic[[v]])) 100 else 1,
           tirz_sd = if (is.logical(analytic[[v]])) NA else sd(x[t1]),
           sema_sd = if (is.logical(analytic[[v]])) NA else sd(x[t0]))
  }
}
table1 <- purrr::map_dfr(t1_vars, make_rows) |>
  mutate(smd_unweighted = NA_real_, smd_weighted = NA_real_)
# attach SMDs from cobalt (names: variable, variable_TRUE or variable_level)
for (i in seq_len(nrow(table1))) {
  key <- c(paste0(table1$variable[i], "_", table1$level[i]), paste0(table1$variable[i], "_TRUE"), table1$variable[i])
  hit <- bal_df[bal_df$covariate %in% key, ]
  if (nrow(hit)) { table1$smd_unweighted[i] <- hit$smd_unweighted[1]; table1$smd_weighted[i] <- hit$smd_weighted[1] }
}
table1 <- bind_rows(
  tibble(variable = "N patients", level = "", type = "n",
         tirz = sum(analytic$treat == 1), sema = sum(analytic$treat == 0),
         tirz_w = sum(analytic$w[analytic$treat == 1]), sema_w = sum(analytic$w[analytic$treat == 0])),
  table1)
readr::write_csv(table1, file.path(dir_tables, "table1_baseline.csv"))
print(table1 |> mutate(across(where(is.numeric), ~ round(.x, 2))), n = Inf, width = 200)
```

**What.** Plot the propensity score overlap (Figure 2) and a Love plot of balance (Figure 3).

**Why.** Overlap checks positivity (each type of patient could have received either drug); the Love plot shows balance at a glance.

```{r ps-figures, fig.height = 9}
p_overlap <- ggplot(analytic, aes(x = ps, fill = drug)) +
  geom_density(alpha = 0.45, colour = NA) +
  scale_fill_manual(values = col_drug, name = NULL) +
  labs(x = "Propensity score (probability of starting tirzepatide)", y = "Density",
       title = "Figure 2. Propensity score overlap") +
  theme(legend.position = "bottom")
ggsave(file.path(dir_figures, "fig2_ps_overlap.png"), p_overlap, width = 7, height = 4.5, dpi = 200, bg = "white")

p_love <- cobalt::love.plot(w_ate, binary = "std", abs = TRUE, thresholds = c(m = 0.1),
                            var.order = "unadjusted", line = FALSE, stars = "none",
                            colors = c("#B5B5B5", "#1F6FB4"), shapes = c("circle", "triangle"),
                            sample.names = c("Unweighted", "IPTW")) +
  labs(title = "Figure 3. Covariate balance before and after weighting")
ggsave(file.path(dir_figures, "fig3_love_plot.png"), p_love, width = 7.5, height = 9, dpi = 200, bg = "white")

summary(analytic$w)
```

# Step 10. Outcome models

**What.** Helper functions: `robust_ci()` (robust sandwich standard errors with `sandwich::vcovHC(type = "HC0")`), `wmean_by()` (weighted means by group) and `binary_effect()` (risk difference from a weighted linear model and risk ratio from a weighted log-link model).

**Why.** Weighted analyses need robust variance; each binary outcome is analysed in the same way.

```{r models-helpers}
analytic <- readRDS(file.path(dir_derived, "analytic_weighted.rds"))
ps_spec  <- readRDS(file.path(dir_derived, "ps_spec.rds"))

robust_ci <- function(fit, term = "treat", exp = FALSE) {
  est <- coef(fit)[[term]]; se <- sqrt(sandwich::vcovHC(fit, type = "HC0")[term, term])
  ci  <- est + c(-1, 1) * qnorm(0.975) * se
  p   <- 2 * pnorm(-abs(est / se))
  if (exp) c(est = exp(est), lo = exp(ci[1]), hi = exp(ci[2]), p = p) else c(est = est, lo = ci[1], hi = ci[2], p = p)
}
wmean_by <- function(y, d = analytic) {
  c(tirz = weighted.mean(y[d$treat == 1], d$w[d$treat == 1]),
    sema = weighted.mean(y[d$treat == 0], d$w[d$treat == 0]))
}
# Binary outcome: weighted proportions, risk difference (linear probability) and risk ratio (log link)
binary_effect <- function(y_name, d = analytic, label = y_name) {
  d$y <- as.numeric(d[[y_name]])
  rd <- robust_ci(lm(y ~ treat, data = d, weights = w))
  rr <- robust_ci(glm(y ~ treat, data = d, weights = w, family = quasipoisson(link = "log")), exp = TRUE)
  m  <- wmean_by(d$y, d)
  tibble(outcome = label, tirz = m[["tirz"]] * 100, sema = m[["sema"]] * 100,
         diff = rd[["est"]] * 100, diff_lo = rd[["lo"]] * 100, diff_hi = rd[["hi"]] * 100,
         ratio = rr[["est"]], ratio_lo = rr[["lo"]], ratio_hi = rr[["hi"]], p = rr[["p"]], measure = "RD (pct points) / RR")
}
```

**What**

- Primary outcome: `binary_effect("persistent")`.
- Time to discontinuation: weighted Kaplan-Meier (`survfit(..., weights = w)`) and weighted Cox model (`coxph(..., robust = TRUE)`); proportional hazards test (`cox.zph()`).
- Restricted mean days on treatment (`rmst()`); Kaplan-Meier figure (Figure 4).

**Why.** Aim 1 asks for both persistence at 12 months and time to discontinuation. RMST is easy to interpret ("days on treatment") and stays valid if hazards are not proportional.

```{r models-persistence}
# Primary outcome: persistence at 12 months
primary <- binary_effect("persistent", label = "Persistent at 12 months (60-day gap)")
print(primary)

# Time to discontinuation: weighted Kaplan-Meier and weighted Cox (robust variance)
km  <- survfit(Surv(time_to_disc, discontinued) ~ drug, data = analytic, weights = w)
cox <- coxph(Surv(time_to_disc, discontinued) ~ treat, data = analytic, weights = w, robust = TRUE)
hr  <- summary(cox)$conf.int
ph_test <- cox.zph(cox)
print(summary(cox)); print(ph_test)

# Restricted mean time on treatment over 365 days (area under each weighted KM curve)
rmst <- function(fit) {
  s <- summary(fit, times = c(0, fit$time), extend = TRUE)
  tapply(seq_along(s$time), s$strata, function(i) {
    t <- c(s$time[i], 365); S <- s$surv[i]
    keep <- t[-length(t)] < 365
    sum(diff(pmin(t, 365))[keep] * S[keep])
  })
}
rmst_km <- rmst(km); print(rmst_km)

p_km <- survminer::ggsurvplot(km, data = analytic, conf.int = TRUE, censor = FALSE,
                              palette = unname(col_drug[c("Semaglutide 2.4 mg", "Tirzepatide")]),
                              legend.labs = c("Semaglutide 2.4 mg", "Tirzepatide"), legend.title = "",
                              xlab = "Days since index", ylab = "Proportion still on index drug",
                              break.time.by = 60, xlim = c(0, 365), ylim = c(0, 1),
                              title = "Figure 4. Time to discontinuation (IPTW-weighted Kaplan-Meier)",
                              ggtheme = theme_minimal(base_size = 11))
ggsave(file.path(dir_figures, "fig4_km_discontinuation.png"), p_km$plot, width = 7.5, height = 5, dpi = 200, bg = "white")
```

**What.** PDC mean difference (weighted linear model), PDC >= 80% and switching (`binary_effect()`), combined into Table 2.

**Why.** Aim 2 (adherence and switching).

```{r models-adherence}
pdc_fit <- lm(pdc ~ treat, data = analytic, weights = w)
pdc_eff <- robust_ci(pdc_fit)
pdc_m   <- wmean_by(analytic$pdc)
table2 <- bind_rows(
  primary,
  tibble(outcome = "Time to discontinuation (hazard ratio)", ratio = hr[1, "exp(coef)"],
         ratio_lo = hr[1, "lower .95"], ratio_hi = hr[1, "upper .95"],
         p = summary(cox)$coefficients[1, "Pr(>|z|)"], measure = "HR"),
  tibble(outcome = "Restricted mean days on treatment (0-365)", tirz = rmst_km[[2]], sema = rmst_km[[1]],
         diff = rmst_km[[2]] - rmst_km[[1]], measure = "Mean difference (days); CI from bootstrap"),
  tibble(outcome = "PDC, mean (%)", tirz = pdc_m[["tirz"]] * 100, sema = pdc_m[["sema"]] * 100,
         diff = pdc_eff[["est"]] * 100, diff_lo = pdc_eff[["lo"]] * 100, diff_hi = pdc_eff[["hi"]] * 100,
         p = pdc_eff[["p"]], measure = "Mean difference (pct points)"),
  binary_effect("pdc80", label = "Adherent (PDC >= 80%)"),
  binary_effect("switched", label = "Switched to the other drug")
)
```

**What.** Any inpatient admission and any ED visit (`binary_effect()`); outpatient and obesity-related visit counts with weighted negative binomial models (`MASS::glm.nb()`), with the Poisson dispersion as a check.

**Why.** Aim 3. Visit counts are overdispersed (variance > mean), which the negative binomial model allows.

```{r models-utilization}
table3_bin <- bind_rows(
  binary_effect("any_ip", label = "Any inpatient admission"),
  binary_effect("any_ed", label = "Any emergency department visit")
)
count_effect <- function(y_name, label) {
  f  <- reformulate("treat", response = y_name)
  nb <- MASS::glm.nb(f, data = analytic, weights = w)
  po <- glm(f, data = analytic, weights = w, family = poisson)
  dispersion <- sum(residuals(po, type = "pearson")^2) / po$df.residual
  rr <- robust_ci(nb, exp = TRUE); m <- wmean_by(analytic[[y_name]])
  list(row = tibble(outcome = label, tirz = m[["tirz"]], sema = m[["sema"]], diff = m[["tirz"]] - m[["sema"]],
                    ratio = rr[["est"]], ratio_lo = rr[["lo"]], ratio_hi = rr[["hi"]], p = rr[["p"]],
                    measure = "Mean difference / rate ratio (negative binomial)"),
       dispersion = dispersion, theta = nb$theta)
}
op  <- count_effect("n_outpatient", "Outpatient visits, mean")
opo <- count_effect("n_outpatient_obesity", "Obesity-related outpatient visits, mean")
```

**What**

- Total and pharmacy costs: weighted gamma GLM with log link (`family = Gamma(link = "log")`) gives the cost ratio.
- Medical costs (some are $0): two-part model, part 1 logistic for any cost and part 2 gamma for positive costs (`twopart_mean()`).
- Weighted mean costs by group (`cost_means`).

**Why.** Aim 4. Costs are skewed and cannot be negative; the gamma-log GLM models the mean directly without retransformation problems, and the two-part model handles zeros.

```{r models-costs}
# Gamma GLM (log link) for strictly positive costs; two-part model for medical costs (zeros allowed)
cost_ratio <- function(y_name) {
  robust_ci(glm(reformulate("treat", response = y_name), data = analytic, weights = w,
                family = Gamma(link = "log")), exp = TRUE)
}
cr_total <- cost_ratio("cost_total"); cr_rx <- cost_ratio("cost_pharmacy")
analytic$any_med_cost <- analytic$cost_medical > 0
part1 <- glm(any_med_cost ~ treat, data = analytic, weights = w, family = quasibinomial)
part2 <- glm(cost_medical ~ treat, data = subset(analytic, cost_medical > 0), weights = w, family = Gamma(link = "log"))
twopart_mean <- function(t) {
  nd <- data.frame(treat = t)
  predict(part1, nd, type = "response") * predict(part2, nd, type = "response")
}
cr_med <- c(est = unname(twopart_mean(1) / twopart_mean(0)))
cost_means <- sapply(c("cost_medical", "cost_pharmacy", "cost_total", "cost_study_drug"), function(v) wmean_by(analytic[[v]]))
print(round(cost_means))
```

**What.** Bootstrap 500 times (`boot_once()`): resample patients, re-estimate the propensity score and weights, and recompute the cost differences and RMST difference; take the 2.5th and 97.5th percentiles as the 95% CI. Assemble and save Tables 2 and 3.

**Why.** Confidence intervals for differences in mean costs should reflect skewness and the uncertainty from estimating the weights; the bootstrap does both.

```{r models-bootstrap}
# Bootstrap (500 replicates): re-estimate the propensity score and weights in each resample,
# then recompute weighted mean cost differences and the restricted mean time on treatment
boot_once <- function(i) {
  d <- analytic[sample.int(nrow(analytic), replace = TRUE), ]
  ps <- fitted(glm(ps_spec$ps_formula, data = d, family = binomial))
  pt <- mean(d$treat)
  d$w <- ifelse(d$treat == 1, pt / ps, (1 - pt) / (1 - ps))
  m <- sapply(c("cost_medical", "cost_pharmacy", "cost_total"), function(v) {
    x <- wmean_by(d[[v]], d); x[["tirz"]] - x[["sema"]] })
  r <- rmst(survfit(Surv(time_to_disc, discontinued) ~ drug, data = d, weights = w))
  c(m, rmst = r[[2]] - r[[1]])
}
set.seed(20261007)
boot <- do.call(rbind, lapply(1:500, boot_once))
boot_ci <- apply(boot, 2, quantile, probs = c(0.025, 0.975))
print(round(boot_ci, 1))
saveRDS(boot, file.path(dir_derived, "bootstrap_replicates.rds"))

table2$diff_lo[table2$outcome == "Restricted mean days on treatment (0-365)"] <- boot_ci[1, "rmst"]
table2$diff_hi[table2$outcome == "Restricted mean days on treatment (0-365)"] <- boot_ci[2, "rmst"]

cost_row <- function(v, label, ratio_ci) {
  has_ci <- length(ratio_ci) > 1
  tibble(outcome = label, tirz = cost_means["tirz", v], sema = cost_means["sema", v],
         diff = cost_means["tirz", v] - cost_means["sema", v],
         diff_lo = boot_ci[1, v], diff_hi = boot_ci[2, v],
         ratio = ratio_ci[["est"]], ratio_lo = if (has_ci) ratio_ci[["lo"]] else NA,
         ratio_hi = if (has_ci) ratio_ci[["hi"]] else NA, p = if (has_ci) ratio_ci[["p"]] else NA,
         measure = "Mean difference $ (bootstrap CI) / cost ratio")
}
table3 <- bind_rows(
  table3_bin, op$row, opo$row,
  cost_row("cost_medical", "Medical cost, mean $ (two-part model)", cr_med),
  cost_row("cost_pharmacy", "Pharmacy cost, mean $ (gamma GLM)", cr_rx),
  cost_row("cost_total", "Total cost, mean $ (gamma GLM)", cr_total),
  tibble(outcome = "  of which study-drug pharmacy cost, mean $", tirz = cost_means["tirz", "cost_study_drug"],
         sema = cost_means["sema", "cost_study_drug"],
         diff = cost_means["tirz", "cost_study_drug"] - cost_means["sema", "cost_study_drug"], measure = "Descriptive")
)
readr::write_csv(table2, file.path(dir_tables, "table2_persistence_adherence.csv"))
readr::write_csv(table3, file.path(dir_tables, "table3_hcru_costs.csv"))
print(table2, width = 200); print(table3, width = 200)
```

**What**

- Proportional hazards global test.
- Modified Park test for the cost variance function (slope near 2 supports gamma).
- Poisson dispersion for visit counts.
- Share of $0 medical costs; Schoenfeld residual plot.

**Why.** Workflow step 9: confirm each model's assumptions before interpreting it, and record any departure.

```{r models-checks}
# Modified Park test on a covariate-adjusted weighted gamma model of total cost:
# slope ~ 2 supports the gamma variance function (0 = Gaussian, 1 = Poisson, 3 = inverse Gaussian)
adj_cost <- glm(update(ps_spec$ps_formula, cost_total ~ treat + .), data = analytic, weights = w,
                family = Gamma(link = "log"))
mu <- fitted(adj_cost)
park <- glm(I((analytic$cost_total - mu)^2) ~ log(mu), family = quasipoisson(link = "log"), weights = analytic$w)
park_slope <- coef(park)[2]

# SAP contingency if the Park slope is far from 2: re-estimate the total-cost ratio under an
# inverse Gaussian variance (slope ~ 3) and a log-normal model with Duan smearing by group
gamma_total <- glm(cost_total ~ treat, data = analytic, weights = w, family = Gamma(link = "log"))
ig_ratio <- robust_ci(glm(cost_total ~ treat, data = analytic, weights = w,
                          family = inverse.gaussian(link = "log"), start = coef(gamma_total)), exp = TRUE)
ln_fit   <- lm(log(cost_total) ~ treat, data = analytic, weights = w)
smear    <- sapply(0:1, function(t) {
  i <- analytic$treat == t; weighted.mean(exp(residuals(ln_fit)[i]), analytic$w[i]) })
ln_ratio <- exp(coef(ln_fit)[["treat"]]) * smear[2] / smear[1]

checks <- c(
  sprintf("Proportional hazards (cox.zph) global p = %.3f", ph_test$table["GLOBAL", "p"]),
  sprintf("Modified Park test slope (total cost) = %.2f (gamma ~ 2)", park_slope),
  sprintf("Total cost ratio: gamma %.3f (%.3f-%.3f); inverse Gaussian %.3f (%.3f-%.3f); log-normal with smearing %.3f",
          cr_total[["est"]], cr_total[["lo"]], cr_total[["hi"]],
          ig_ratio[["est"]], ig_ratio[["lo"]], ig_ratio[["hi"]], ln_ratio),
  sprintf("Poisson dispersion, outpatient visits = %.2f (>1 = overdispersion; negative binomial theta = %.2f)",
          op$dispersion, op$theta),
  sprintf("Poisson dispersion, obesity-related visits = %.2f", opo$dispersion),
  sprintf("Share of patients with $0 medical cost = %.1f%%", 100 * mean(analytic$cost_medical == 0))
)
writeLines(checks, file.path(dir_tables, "model_checks.txt")); cat(checks, sep = "\n")
png(file.path(dir_figures, "figS1_schoenfeld.png"), width = 1400, height = 900, res = 200)
plot(ph_test, resid = FALSE, main = "Scaled Schoenfeld residuals: tirzepatide vs semaglutide"); abline(h = coef(cox), lty = 2, col = "red")
dev.off()
```

**What.** Plot weighted cost distributions on a log scale (Figure 5).

**Why.** It shows the skewness that justified the gamma and two-part models.

```{r models-cost-figure, fig.height = 7}
cost_long <- analytic |>
  select(drug, w, Medical = cost_medical, Pharmacy = cost_pharmacy, Total = cost_total) |>
  tidyr::pivot_longer(Medical:Total, names_to = "component", values_to = "cost")
p_cost <- ggplot(cost_long, aes(x = cost + 1, fill = drug, weight = w)) +
  geom_density(alpha = 0.45, colour = NA) +
  scale_x_log10(labels = scales::dollar) +
  facet_wrap(~ component, ncol = 1, scales = "free_y") +
  scale_fill_manual(values = col_drug, name = NULL) +
  labs(x = "12-month allowed cost + $1 (log scale)", y = "Weighted density",
       title = "Figure 5. Distribution of 12-month costs by drug (IPTW)") +
  theme(legend.position = "bottom")
ggsave(file.path(dir_figures, "fig5_cost_distributions.png"), p_cost, width = 7, height = 7, dpi = 200, bg = "white")
```

# Step 11. Subgroup, sensitivity and bias analyses

**What.** Helpers: `persist_effect()` (weighted RD and RR for any data, outcome and weight column) and `reweight()` (re-estimates weights when the population changes).

**Why.** Every sensitivity analysis is computed the same way, so only the planned change differs from the primary analysis.

```{r sens-helpers}
analytic <- readRDS(file.path(dir_derived, "analytic_weighted.rds"))
ps_spec  <- readRDS(file.path(dir_derived, "ps_spec.rds"))

robust_ci <- function(fit, term = "treat", exp = FALSE) {
  est <- coef(fit)[[term]]; se <- sqrt(sandwich::vcovHC(fit, type = "HC0")[term, term])
  ci  <- est + c(-1, 1) * qnorm(0.975) * se
  if (exp) c(est = exp(est), lo = exp(ci[1]), hi = exp(ci[2])) else c(est = est, lo = ci[1], hi = ci[2])
}
# Weighted persistence comparison in data `d` with weights column `w`
persist_effect <- function(d, label, y = "persistent", wcol = "w") {
  d$y <- as.numeric(d[[y]]); d$wt <- d[[wcol]]
  rd_ci <- robust_ci(lm(y ~ treat, data = d, weights = wt))
  rr_ci <- robust_ci(glm(y ~ treat, data = d, weights = wt, family = quasipoisson(link = "log")), exp = TRUE)
  tibble(analysis = label, n = nrow(d),
         tirz = 100 * weighted.mean(d$y[d$treat == 1], d$wt[d$treat == 1]),
         sema = 100 * weighted.mean(d$y[d$treat == 0], d$wt[d$treat == 0]),
         rd = 100 * rd_ci[["est"]], rd_lo = 100 * rd_ci[["lo"]], rd_hi = 100 * rd_ci[["hi"]],
         rr = rr_ci[["est"]], rr_lo = rr_ci[["lo"]], rr_hi = rr_ci[["hi"]])
}
# Re-estimate stabilized ATE weights within a subset (the population changes)
reweight <- function(d, drop = character()) {
  f <- reformulate(setdiff(ps_spec$covs, drop), response = "treat")
  d$w <- WeightIt::weightit(f, data = droplevels(d), method = "glm", estimand = "ATE", stabilize = TRUE)$weights
  d
}
```

**What.** For each prespecified subgroup (obesity class, sex, age group, CDHP), estimate RD and RR within each level and test the treatment-by-subgroup interaction with a robust multi-degree-of-freedom Wald test (`t(b) %*% solve(V) %*% b`).

**Why.** Aim 5. Interaction tests guard against over-reading chance differences between subgroups.

```{r subgroups}
subgroup_vars <- c(obesity_class = "Obesity class", sex = "Sex", age_group = "Age group", cdhp = "Plan type")
subgroups <- purrr::imap_dfr(subgroup_vars, function(lab, v) {
  # interaction test: weighted log-link model with treat x subgroup, robust Wald test (multi-df)
  fit <- glm(reformulate(c("treat", v, paste0("treat:", v)), response = "as.numeric(persistent)"),
             data = analytic, weights = w, family = quasipoisson(link = "log"))
  ix  <- grep("^treat:", names(coef(fit)))
  V   <- sandwich::vcovHC(fit, type = "HC0")[ix, ix, drop = FALSE]
  b   <- coef(fit)[ix]
  p_int <- pchisq(as.numeric(t(b) %*% solve(V) %*% b), df = length(ix), lower.tail = FALSE)
  purrr::map_dfr(levels(analytic[[v]]), function(l) {
    persist_effect(analytic[analytic[[v]] == l, ], label = l)
  }) |> mutate(subgroup = lab, p_interaction = p_int, .before = 1)
})
readr::write_csv(subgroups, file.path(dir_tables, "table4_subgroups.csv"))
print(subgroups, width = 200)
```

**What.** Sensitivity analyses from the SAP:

- S1-S3: alternative persistence definitions;
- S4: 1:1 PS matching (`MatchIt::matchit(caliper = 0.2)`);
- S5: trimmed weights;
- S6: doubly robust g-computation (`marginaleffects::avg_comparisons()`);
- S7: excluding metformin users;
- S8: recorded obesity class only;
- S11: negative control outcome.

**Why.** Aim 6. If conclusions stay the same across definitions and methods, they are less likely to be artefacts of a single analytic choice.

```{r sensitivity}
primary <- persist_effect(analytic, "Primary: IPTW, 60-day gap")

# S4: 1:1 nearest-neighbour matching on the logit PS, caliper 0.2 SD (ATT)
m_out <- MatchIt::matchit(ps_spec$ps_formula, data = analytic, method = "nearest",
                          distance = "glm", link = "linear.logit", caliper = 0.2, estimand = "ATT")
matched <- MatchIt::match.data(m_out)
bal_m <- cobalt::bal.tab(m_out, binary = "std")$Balance
max_smd_matched <- max(abs(bal_m[rownames(bal_m) != "distance", "Diff.Adj"]), na.rm = TRUE)  # covariates only

# S5: weights truncated at the 1st and 99th percentiles
trimmed <- analytic |> mutate(w = pmin(pmax(w, quantile(w, 0.01)), quantile(w, 0.99)))

# S6: doubly robust g-computation (weighted outcome model with all covariates)
dr_fit <- glm(update(ps_spec$ps_formula, as.numeric(persistent) ~ treat + .), data = analytic,
              weights = w, family = quasibinomial)
dr_rd <- marginaleffects::avg_comparisons(dr_fit, variables = "treat", wts = "w", vcov = "HC0")
dr_rr <- marginaleffects::avg_comparisons(dr_fit, variables = "treat", wts = "w", vcov = "HC0",
                                          comparison = "lnratioavg", transform = exp)

sensitivity <- bind_rows(
  primary,
  persist_effect(analytic, "S1: 30-day gap", y = "persistent_g30"),
  persist_effect(analytic, "S2: 90-day gap", y = "persistent_g90"),
  persist_effect(analytic, "S3: Either drug (switching allowed)", y = "persistent_class"),
  persist_effect(matched, "S4: 1:1 PS matching (ATT)", wcol = "weights"),
  persist_effect(trimmed, "S5: Weights trimmed at 1st/99th percentile"),
  tibble(analysis = "S6: Doubly robust (IPTW + outcome model)", n = nrow(analytic),
         rd = 100 * dr_rd$estimate, rd_lo = 100 * dr_rd$conf.low, rd_hi = 100 * dr_rd$conf.high,
         rr = dr_rr$estimate, rr_lo = dr_rr$conf.low, rr_hi = dr_rr$conf.high),
  persist_effect(reweight(filter(analytic, !metformin)), "S7: Excluding baseline metformin users"),
  persist_effect(reweight(filter(analytic, obesity_class != "Unknown")), "S8: Recorded obesity class only"),
  persist_effect(analytic, "S11: Negative control outcome - cancer screening", y = "neg_control")
)
print(sensitivity, width = 200)
message("Matched pairs: ", sum(matched$treat == 1), "; largest covariate SMD after matching: ", round(max_smd_matched, 3))
```

**What.** S9: winsorize costs at the 99th percentile (`wins()`) and recompute weighted mean differences.

**Why.** It shows whether a few very expensive patients drive the cost results.

```{r sensitivity-costs}
# S9: costs winsorized at the 99th percentile (all patients pooled)
wins <- function(x) pmin(x, quantile(x, 0.99))
cost_sens <- purrr::map_dfr(c("cost_medical", "cost_pharmacy", "cost_total"), function(v) {
  x <- analytic[[v]]; xw <- wins(x); t1 <- analytic$treat == 1
  tibble(outcome = v,
         diff_raw  = weighted.mean(x[t1], analytic$w[t1]) - weighted.mean(x[!t1], analytic$w[!t1]),
         diff_wins = weighted.mean(xw[t1], analytic$w[t1]) - weighted.mean(xw[!t1], analytic$w[!t1]),
         ratio_wins = weighted.mean(xw[t1], analytic$w[t1]) / weighted.mean(xw[!t1], analytic$w[!t1]))
})
readr::write_csv(cost_sens, file.path(dir_tables, "table5b_cost_sensitivity.csv"))
print(cost_sens)
```

**What.** S10: E-values for the primary risk ratio (`EValue::evalues.RR()`) and hazard ratio (`evalues.HR(rare = FALSE)`).

**Why.** The E-value (VanderWeele & Ding 2017) is the minimum strength of association, on the risk ratio scale, that an unmeasured confounder would need with both drug choice and persistence to explain away the result. It is reported for the point estimate and for the confidence limit closest to 1.

```{r evalues}
# S10: E-values (VanderWeele & Ding, Ann Intern Med 2017): the minimum strength of unmeasured
# confounding, on the RR scale, needed to explain away the estimate and its confidence limit.
# Persistence is common (>15%), so the HR is converted (rare = FALSE).
cox <- coxph(Surv(time_to_disc, discontinued) ~ treat, data = analytic, weights = w, robust = TRUE)
hr  <- summary(cox)$conf.int
ev_rr <- EValue::evalues.RR(primary$rr, primary$rr_lo, primary$rr_hi)
ev_hr <- EValue::evalues.HR(hr[1, 1], hr[1, 3], hr[1, 4], rare = FALSE)
evalues <- tibble(
  estimate = c("RR persistent at 12 months", "HR discontinuation"),
  point = c(primary$rr, hr[1, 1]), lo = c(primary$rr_lo, hr[1, 3]), hi = c(primary$rr_hi, hr[1, 4]),
  evalue_point = c(ev_rr["E-values", "point"], ev_hr["E-values", "point"]),
  evalue_ci    = c(ev_rr["E-values", "lower"], ev_hr["E-values", "upper"]))
readr::write_csv(evalues, file.path(dir_tables, "evalues.csv"))
print(evalues)
readr::write_csv(sensitivity, file.path(dir_tables, "table5_sensitivity.csv"))
```

**What.** Forest plots of the subgroup (Figure 6) and sensitivity (Figure 7) results.

**Why.** It summarises robustness visually for the paper.

```{r sens-figures, fig.height = 5}
forest <- function(d, ylab_col, title, xlab = "Risk ratio for 12-month persistence (95% CI)") {
  ggplot(d, aes(x = rr, xmin = rr_lo, xmax = rr_hi, y = forcats::fct_rev(forcats::fct_inorder(.data[[ylab_col]])))) +
    geom_vline(xintercept = 1, linetype = 2, colour = "grey50") +
    geom_pointrange(colour = "#1F6FB4", size = 0.35) +
    labs(x = xlab, y = NULL, title = title)
}
sg_plot <- subgroups |> mutate(label = paste0(subgroup, ": ", analysis, "  (p-int ", signif(p_interaction, 2), ")"))
ggsave(file.path(dir_figures, "fig6_subgroups.png"),
       forest(sg_plot, "label", "Figure 6. Persistence by prespecified subgroup (IPTW)"),
       width = 8, height = 4.5, dpi = 200, bg = "white")
ggsave(file.path(dir_figures, "fig7_sensitivity.png"),
       forest(sensitivity, "analysis", "Figure 7. Sensitivity analyses (risk ratio, tirzepatide vs semaglutide)",
              xlab = "Risk ratio (95% CI); S11 is the negative control outcome"),
       width = 8.5, height = 4.5, dpi = 200, bg = "white")
```

# Session information

```{r session-info}
sessionInfo()
```
