---
title: "Tirzepatide vs Semaglutide 2.4 mg: Persistence, Utilization and Costs"
subtitle: "Annotated analysis of a synthetic commercial claims extract (target trial emulation)"
author: "Maxwell K. Nyarko"
date: "October 2026"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: false
    df_print: paged
---

```{r knitr-options, include = FALSE}
# Run every chunk from the project folder; show code and results; do not stop on warnings
knitr::opts_knit$set(root.dir = here::here())
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 8, fig.height = 5)
```

# How to use this file

1. Open `HEOR_Tirzepatide_Project.Rproj` in RStudio (this sets the working folder).
2. Open this file and either click **Knit** (runs everything and makes an HTML report) or run the chunks one at a time from top to bottom with the green arrow on each chunk (Ctrl/Cmd + Shift + Enter).
3. The first chunk installs any missing packages, which can take a few minutes the first time. A full run takes about 5-10 minutes; the bootstrap in Step 9.6 is the slowest part.

Every step below has the same layout: **Step**, **What** (each key action and the code line that does it), **Why**, then the R code. Each step saves its results to `04_data_derived/` or `06_output/`, so you can stop and restart at any step after the setup chunk.

**The data are fully synthetic.** Results show how the methods work and are not evidence about either drug.

The same code is also saved as numbered scripts in `05_code/` (00 to 10); this file is assembled from those scripts by `05_code/98_build_rmd.R`.

# Step 0. Set up R

**What**

- Install and load the packages for data handling, weighting, models and figures: `install.packages(to_install)` and `lapply(pkgs, library, ...)`.
- Make `select()` refer to dplyr (MASS also has a `select()`): `select <- dplyr::select`.

**Why.** Every later step uses these packages. Loading them once at the start makes the run reproducible and avoids errors halfway through.

{{chunk:setup-packages}}

**What**

- Define folder paths with `here::here()` so the code works on any computer.
- Store the study calendar from the protocol: identification window (`index_start`, `index_end`), 365-day baseline and follow-up, the 60-day persistence gap (`gap_days`), and the 30-day allowed enrollment gap.
- Fix the random seed with `set.seed(20261007)` so the bootstrap gives the same answer each time.

**Why.** Keeping all design constants in one place means a design change is made once, and readers can check the code against the SAP.

{{chunk:setup-paths}}

# Step 1. Study design: the DAG (workflow step 3)

**What**

- Write the assumed causal structure as a DAG with `dagitty::dagitty()`: one for persistence (`dag_persist`) and one for costs (`dag_cost`).
- Mark the exposure (`Drug`) and outcome, and list unmeasured nodes (`SES_U`, `Preference_U`).

**Why.** The DAG makes our assumptions explicit and tells us which variables to adjust for (confounders) and which to leave alone (mediators such as GI side effects or the drug's copay).

{{chunk:dag-define}}

**What**

- Ask dagitty for the minimal sufficient adjustment set: `adjustmentSets(dag_persist, type = "minimal")`.
- List the descendants of `Drug`, which must not be adjusted for: `descendants(dag_persist, "Drug")`.

**Why.** This turns the DAG into a concrete covariate list for the propensity score model (Step 8).

{{chunk:dag-adjust}}

**What.** Plot both DAGs with nodes coloured by role (`plot_dag()`) and save them to `06_output/figures/`.

**Why.** A picture of the assumptions goes into the DAG document and the paper.

{{chunk:dag-plot, fig.height = 7}}

# Step 2. Import the raw extract (workflow step 7 begins)

**What**

- Declare column types so all codes are read as text (`rx_types`, `med_types`; e.g. `ndc = col_character()`).
- Read and stack the yearly gzip files with `read_stack()`.
- Read the members, enrollment and reference tables with `readr::read_csv()`.

**Why.** Reading NDC, ICD-10, CPT and place-of-service codes as numbers would drop leading zeros and silently break code matching. Nothing is cleaned here, so the raw data stay exactly as delivered.

{{chunk:import-raw}}

**What.** Print row and column counts and parsing problems for each table, then save R copies with `saveRDS()`.

**Why.** A structural check right after import catches truncated files or type problems before they cause trouble later.

{{chunk:import-check}}

# Step 3. Data quality review

**What**

- Reload the raw tables (`readRDS()`).
- Create a log with one row per check, using `log_check(table, check, n_affected, decision)`.

**Why.** Vendor claims always contain errors. Finding them before cleaning, and writing down a decision for each, makes the cleaning transparent and reproducible (workflow step 7: "log every data-cleaning decision").

{{chunk:dq-load}}

**What.** Members: count exact duplicates (`duplicated()`), IDs with conflicting demographics (`count(member_id) |> filter(n > 1)`), missing birth year or state, and unknown sex.

**Why.** A member ID should identify one person. Conflicting records mean we cannot know the person's sex or age.

{{chunk:dq-members}}

**What.** Enrollment: count duplicate segments, overlapping segments (`eligibility_start <= prev_end`), gaps (`eligibility_start > prev_end + 1`), and segments without drug or medical coverage.

**Why.** Continuous enrollment is an eligibility criterion, and pharmacy claims exist only while drug coverage is active.

{{chunk:dq-enrollment}}

**What.** Pharmacy: count the three NDC formats (`nchar(pharmacy$ndc)`), duplicates, reversals (`claim_status == "R"`), missing allowed amounts, and implausible days supply (< 3 or > 90).

**Why.** NDCs must match the lookup to identify the study drugs; reversals are claims that were never dispensed; days supply drives persistence and PDC.

{{chunk:dq-pharmacy}}

**What.** Medical: count duplicates, lower-case and dotted ICD-10 codes, place of service with a lost zero, missing or negative allowed amounts, very large amounts, and obesity-class codes used before their effective date.

**Why.** Diagnosis codes define obesity, diabetes and comorbidities; cost fields define the economic outcomes.

{{chunk:dq-medical}}

**What.** Combine the checks with `bind_rows(dq)` and save `06_output/tables/data_quality_log.csv`.

**Why.** The log is a deliverable and is summarized in the data description document.

{{chunk:dq-save}}

# Step 4. Clean members and enrollment

**What**

- Drop exact duplicate member rows (`distinct()`) and flag IDs that still have more than one row (`demo_conflict = n() > 1`).
- Add census region from the state lookup (`left_join(... state_census_region.csv)`), with "Unknown" when state is missing.

**Why.** One clean row per member is needed before joining; flagged members are excluded at cohort step 4.

{{chunk:clean-members}}

**What**

- Keep segments with both medical and pharmacy coverage (`filter(medical_coverage == "Y", rx_coverage == "Y")`).
- Merge overlapping or nearly adjacent segments into continuous spans: `cummax()` tracks the latest end date, and a new span starts only when the next segment begins more than 30 days later (`new_span`).

**Why.** Continuous enrollment (365 days before and after index) guarantees we see all claims in the baseline and follow-up windows, so a missing claim means no care, not missing data.

{{chunk:clean-enrollment}}

# Step 5. Clean pharmacy and medical claims

**What**

- Convert every NDC to 11 digits with `normalize_ndc()` (remove hyphens, pad the labeler, restore lost leading zeros).
- Remove reversed claims and the paid claims they reverse (`!rx_claim_id %in% reversed_ids`), and keep one row per claim (`distinct(rx_claim_id, ...)`).
- Fill missing allowed amounts as paid + out-of-pocket (`coalesce()`), and add drug names from the lookup (`left_join(ndc_lookup ...)`).
- Replace implausible days supply with the most common value for the same NDC and quantity (`ds_mode`, `days_supply_fixed`).
- Check that every NDC matched (`stopifnot()`).

**Why.** Without these fixes some Zepbound and Wegovy fills would be missed, reversed fills would count as treatment, and a "999-day" supply would make a patient look persistent for years.

{{chunk:clean-pharmacy}}

**What**

- Standardize diagnosis codes (`clean_dx()`: upper case, no dot) and pad place of service to two characters (`str_pad()`).
- Remove duplicate lines (`distinct()`, then one row per `claim_id` + `line_num`).
- Back-calculate missing allowed amounts from paid amounts using the median paid/allowed ratio by claim type (`paid_ratio`).

**Why.** Consistent codes make the code lists work, and duplicate lines would double-count utilization and costs.

{{chunk:clean-medical}}

# Step 6. Build the cohort (with a flow diagram)

**What**

- Load the clean tables and build a long diagnosis table, one row per member, date and code (`pivot_longer(dx1:dx5)`).

**Why.** Searching one column is simpler and faster than searching five.

{{chunk:cohort-load}}

**What**

- Define the code lists: study products, all GLP-1 RA products, obesity (`dx_obesity`), diabetes (`dx_diabetes`), contraindications (`dx_contraind`) and non-metformin antidiabetic drug classes.
- Write a helper, `has_dx()`, that returns TRUE if a member has a matching code between two dates.

**Why.** Code lists written in one place are easy to review against the SAP and reuse.

{{chunk:cohort-codes}}

**What**

- Find each member's first Zepbound or Wegovy fill in January 2024 to June 2025: this is the index date and sets the treatment group (`summarise(index_date = first(fill_date))`).
- Record the counts at each step with `add_step()`.

**Why.** In the target trial emulation, the first fill is "time zero": eligibility, treatment assignment and the start of follow-up all happen on the same day, which avoids immortal time bias.

{{chunk:cohort-index}}

**What.** Apply the criteria in SAP order, recording counts after each `filter()` / `semi_join()` / `anti_join()`:

1. one study drug on the index date;
2. age 18-64;
3. known sex and no conflicting records;
4. continuous enrollment (`span_start <= index_date - 365`, `span_end >= index_date + 364`);
5. no GLP-1 RA fill in the 365-day baseline (new user);
6. obesity diagnosis;
7. no diabetes (no E10/E11 code and no non-metformin antidiabetic fill);
8. no thyroid cancer / MEN 2 code.

**Why.** Each criterion puts into practice an element of the PECO question and the target trial. The new-user washout makes the two groups comparable at the start of treatment.

{{chunk:cohort-criteria}}

**What.** Add plan type and employer group from the enrollment segment covering the index date (`eligibility_start <= index_date`, `eligibility_end >= index_date`), add the index quarter, and save `cohort.rds`.

**Why.** Plan type is a confounder (benefit design and cost sharing). Index quarter captures calendar-time changes in supply and coverage.

{{chunk:cohort-plan}}

**What.** Save the attrition table and draw the flow diagram (Figure 1) with ggplot2.

**Why.** The flow diagram is required by RECORD-PE and shows readers how the study population was reached.

{{chunk:cohort-attrition, fig.height = 8}}

# Step 7. Baseline covariates

**What.** Load the cohort and claims, and compute each claim's day relative to the index date (`day = fill_date - index_date`).

**Why.** Working in "days since index" makes every time window a simple filter (e.g., baseline = `day >= -365 & day <= -1`).

{{chunk:cov-load}}

**What**

- Define encounters once, one per member per day: inpatient admissions (`claim_type == "INST"` with an admission date), ED visits (place of service 23, revenue code 0450 or CPT 99281-99285), outpatient E/M visits, and obesity-related outpatient visits (E66/Z68 on the claim).
- Sum medical costs by claim and floor each claim at $0 (`max(0, sum(allowed_amt))`).

**Why.** The same definitions are used for baseline covariates (here) and follow-up outcomes (Step 8), so the two are measured consistently.

{{chunk:cov-encounters}}

**What**

- Flag comorbidities from codes in [index - 365, index] with `dx_flag()` (e.g., hypertension `^I1[0-6]`, sleep apnea `^G4733`).
- Assign obesity class from the most recent BMI or class code (`case_when()`; Class 3 = BMI >= 40, E66.01, E66.2 or E66.813).

**Why.** These are the confounders in the DAG: they influence which drug is prescribed and the outcomes.

{{chunk:cov-diagnoses}}

**What.** Baseline utilization and costs: counts of each encounter type (`count(member_id, type)`), baseline medical and pharmacy costs, number of distinct generic drugs, endocrinologist visit, and prior bariatric surgery.

**Why.** Prior healthcare use is the best available proxy for overall health and care-seeking, which affect both drug choice and future costs.

{{chunk:cov-utilization}}

**What.** Baseline medication flags by therapeutic class with `rx_flag()` (other anti-obesity drugs, metformin, antihypertensives, statins, antidepressants).

**Why.** Medication use marks comorbidity severity and treatment history.

{{chunk:cov-medications}}

**What.** Join everything into one row per patient; create factors (obesity class, age group, CDHP vs non-CDHP), any baseline inpatient/ED use, and `log_base_cost = log1p(base_total_cost)`; save `covariates.rds`.

**Why.** The log of baseline cost reduces the influence of a few very expensive patients in the propensity score model.

{{chunk:cov-assemble}}

# Step 8. Outcomes over the 365-day follow-up

**What.** Load data; keep follow-up fills (days 0-364) of either molecule, labelled by molecule (Zepbound/Mounjaro = tirzepatide; Wegovy/Ozempic/Rybelsus = semaglutide).

**Why.** Persistence is measured on the index molecule, and fills of the other molecule mark switching.

{{chunk:out-load}}

**What.** Write the persistence function `persistence(day, ds, gap)`:

- `start`/`end` lay the fills end to end, carrying early refills forward;
- `gaps_before` and `tail_gap` measure days without supply;
- the first gap longer than `gap` days is a discontinuation (`disc_day`);
- `pdc` = days covered / 365.

**Why.** This is the standard claims definition of persistence (no refill gap > 60 days) and adherence (proportion of days covered), as used by Marshall 2026. Writing it as a function lets us change the gap for the sensitivity analyses.

{{chunk:out-persistence-function}}

**What.** Apply the function to each patient (`group_by(member_id) |> reframe(persistence(...))`) with gaps of 60 (primary), 30 and 90 days, and on either molecule (`p_class`). Flag switching to the other molecule (`switching`).

**Why.** The primary outcome and sensitivity analyses S1-S3 come from the same function, so only the definition changes.

{{chunk:out-persistence}}

**What.** Count follow-up encounters by type, and sum 12-month medical, pharmacy, study-drug and out-of-pocket costs. Flag the negative control outcome (colonoscopy 45378 or screening mammogram 77067).

**Why.** These are the Aim 3 and 4 outcomes. The negative control outcome checks for residual confounding by healthcare-seeking behaviour.

{{chunk:out-utilization-costs}}

**What.** Join all outcomes, fill zeros for patients with no encounters (`coalesce(.x, 0)`), create binary outcomes (`pdc80`, `any_ip`, `any_ed`) and total cost; save `analytic.rds`; print a crude (unweighted) summary by drug.

**Why.** One analytic dataset with one row per patient feeds every model. The crude summary is a first look before adjustment.

{{chunk:out-assemble}}

# Step 9. Table 1, propensity score and weighting

**What.** Load the analytic file and write the propensity score formula from the DAG adjustment set (`covs`, `reformulate(covs, response = "treat")`).

**Why.** The PS model includes the confounders and nothing measured after the index date.

{{chunk:ps-formula}}

**What.** Fit a logistic propensity score model and compute stabilized ATE weights with `WeightIt::weightit(..., method = "glm", estimand = "ATE", stabilize = TRUE)`; save the weights.

**Why.** Inverse probability of treatment weighting creates a pseudo-population in which measured confounders are balanced between the drugs, as in a randomized trial. Stabilization keeps the weights near 1.

{{chunk:ps-weights}}

**What.** Check balance with `cobalt::bal.tab()`: standardized mean differences (SMD) and variance ratios before and after weighting; report the largest SMD and the effective sample size.

**Why.** Weighting only works if balance is achieved; the SAP target is |SMD| < 0.10 for every covariate.

{{chunk:ps-balance}}

**What.** Build Table 1: unweighted and weighted means or percentages by group, with SMDs (`make_rows()`); save to CSV.

**Why.** Table 1 describes the study population and shows the reader that the weighted groups are comparable.

{{chunk:table1}}

**What.** Plot the propensity score overlap (Figure 2) and a Love plot of balance (Figure 3).

**Why.** Overlap checks positivity (each type of patient could have received either drug); the Love plot shows balance at a glance.

{{chunk:ps-figures, fig.height = 9}}

# Step 10. Outcome models

**What.** Helper functions: `robust_ci()` (robust sandwich standard errors with `sandwich::vcovHC(type = "HC0")`), `wmean_by()` (weighted means by group) and `binary_effect()` (risk difference from a weighted linear model and risk ratio from a weighted log-link model).

**Why.** Weighted analyses need robust variance; each binary outcome is analysed in the same way.

{{chunk:models-helpers}}

**What**

- Primary outcome: `binary_effect("persistent")`.
- Time to discontinuation: weighted Kaplan-Meier (`survfit(..., weights = w)`) and weighted Cox model (`coxph(..., robust = TRUE)`); proportional hazards test (`cox.zph()`).
- Restricted mean days on treatment (`rmst()`); Kaplan-Meier figure (Figure 4).

**Why.** Aim 1 asks for both persistence at 12 months and time to discontinuation. RMST is easy to interpret ("days on treatment") and stays valid if hazards are not proportional.

{{chunk:models-persistence}}

**What.** PDC mean difference (weighted linear model), PDC >= 80% and switching (`binary_effect()`), combined into Table 2.

**Why.** Aim 2 (adherence and switching).

{{chunk:models-adherence}}

**What.** Any inpatient admission and any ED visit (`binary_effect()`); outpatient and obesity-related visit counts with weighted negative binomial models (`MASS::glm.nb()`), with the Poisson dispersion as a check.

**Why.** Aim 3. Visit counts are overdispersed (variance > mean), which the negative binomial model allows.

{{chunk:models-utilization}}

**What**

- Total and pharmacy costs: weighted gamma GLM with log link (`family = Gamma(link = "log")`) gives the cost ratio.
- Medical costs (some are $0): two-part model, part 1 logistic for any cost and part 2 gamma for positive costs (`twopart_mean()`).
- Weighted mean costs by group (`cost_means`).

**Why.** Aim 4. Costs are skewed and cannot be negative; the gamma-log GLM models the mean directly without retransformation problems, and the two-part model handles zeros.

{{chunk:models-costs}}

**What.** Bootstrap 500 times (`boot_once()`): resample patients, re-estimate the propensity score and weights, and recompute the cost differences and RMST difference; take the 2.5th and 97.5th percentiles as the 95% CI. Assemble and save Tables 2 and 3.

**Why.** Confidence intervals for differences in mean costs should reflect skewness and the uncertainty from estimating the weights; the bootstrap does both.

{{chunk:models-bootstrap}}

**What**

- Proportional hazards global test.
- Modified Park test for the cost variance function (slope near 2 supports gamma).
- Poisson dispersion for visit counts.
- Share of $0 medical costs; Schoenfeld residual plot.

**Why.** Workflow step 9: confirm each model's assumptions before interpreting it, and record any departure.

{{chunk:models-checks}}

**What.** Plot weighted cost distributions on a log scale (Figure 5).

**Why.** It shows the skewness that justified the gamma and two-part models.

{{chunk:models-cost-figure, fig.height = 7}}

# Step 11. Subgroup, sensitivity and bias analyses

**What.** Helpers: `persist_effect()` (weighted RD and RR for any data, outcome and weight column) and `reweight()` (re-estimates weights when the population changes).

**Why.** Every sensitivity analysis is computed the same way, so only the planned change differs from the primary analysis.

{{chunk:sens-helpers}}

**What.** For each prespecified subgroup (obesity class, sex, age group, CDHP), estimate RD and RR within each level and test the treatment-by-subgroup interaction with a robust multi-degree-of-freedom Wald test (`t(b) %*% solve(V) %*% b`).

**Why.** Aim 5. Interaction tests guard against over-reading chance differences between subgroups.

{{chunk:subgroups}}

**What.** Sensitivity analyses from the SAP:

- S1-S3: alternative persistence definitions;
- S4: 1:1 PS matching (`MatchIt::matchit(caliper = 0.2)`);
- S5: trimmed weights;
- S6: doubly robust g-computation (`marginaleffects::avg_comparisons()`);
- S7: excluding metformin users;
- S8: recorded obesity class only;
- S11: negative control outcome.

**Why.** Aim 6. If conclusions stay the same across definitions and methods, they are less likely to be artefacts of a single analytic choice.

{{chunk:sensitivity}}

**What.** S9: winsorize costs at the 99th percentile (`wins()`) and recompute weighted mean differences.

**Why.** It shows whether a few very expensive patients drive the cost results.

{{chunk:sensitivity-costs}}

**What.** S10: E-values for the primary risk ratio (`EValue::evalues.RR()`) and hazard ratio (`evalues.HR(rare = FALSE)`).

**Why.** The E-value (VanderWeele & Ding 2017) is the minimum strength of association, on the risk ratio scale, that an unmeasured confounder would need with both drug choice and persistence to explain away the result. It is reported for the point estimate and for the confidence limit closest to 1.

{{chunk:evalues}}

**What.** Forest plots of the subgroup (Figure 6) and sensitivity (Figure 7) results.

**Why.** It summarises robustness visually for the paper.

{{chunk:sens-figures, fig.height = 5}}

# Session information

```{r session-info}
sessionInfo()
```
